### 2021-08-21 FormalMethodologyWriteup.Rmd
Adding files from R

---
title: "Formal Writeup"
author: "Paul Harmon"
date: "7/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("OutlierCompFunctions.R")
library(ggplot2)
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(janitor)
library(magrittr)
library(knitr)
library(Rtsne)
library(robustHD)
library(GGally)
library(stringr)
library(plotly)
library(R.utils)
library(vegan)
library(mnormt)

Generate_Holdout_Maps <- function(data, j, ...){

  #hold out a single row of the data (in the jth row)
  #note - have to remove, as t-SNE doesn't like NAs
  data_holdout <- data[-j,]

  #calculate distances FIRST
  data_dist <- dist(data_holdout)

  #### Creates Maps (based on several methods)

  #generate the classical MDS
  mds1 <- cmdscale(data_dist, k = 2)

  #generate the t-SNE map and store x,y in map list
  x <- Rtsne(data_dist, perplexity = perplexity, pca = pca_options, is_distance = TRUE)

  return(list(mds = mds1, tsne = x$Y))
}

perplexity = 10

```

# Introduction

Consider the problem of visualizing multivariate (or potentially high-dimensional) data.  In many cases, it is necessary to generate a 2 or 3-D mapping of some higher-dimensional space, projecting the high-dimensional data into an ordination that can be printed on paper. There are many methods that exist to accomplish this task, ranging from tools such as classicial multidimensional scaling (MDS) to more modern tools like t-distributed Stochastic Neighbor Embedding.

Unlike regression or classification, most tools for dimension reduction and higher-dimensional data visualization are unsupervised, meaning that they do not take advantage of some marked response variable; rather, they capitalize on the signal present in a set of features and produce a data-driven result.

Despite those differences, tools like t-SNE and MDS are succeptible to abberant data values, much in the same way that a regression model might be succeptible to outliers and influential points.  In a regression model, the estimated coefficients or predictions might be affected in meaningful fashion by the inclusion or exclusion of a single point - such points are referred to as influential (Cook, 1977).  A suite of tools have been developed to identify and measure the impact of influential points on the results of regression models.

## Goal

The goal of this method is to develop a diagnostic that can be used to help identify influential points and potentially quantify how much they impact the resulting mapping, irrespective of how that mapping was created.  Additionally, this method can be used to assess the sensitivity of each method to influential points and the impact that sensititivy might have on how an end user might want to interpret the resulting map produced.  Note that while robustness to influential points might at first seem like a positive trait for an ordination method, masking distinctly different observations in a mapping may have negative consequences as well.

This document overviews the statistical methodology we have developed to identify influential points that, when excluded from a dataset, can meaningfully impact the shape of an ordination created from data. We overview our methodology step by step.

# Methods

This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE. In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.

Our Permanova-based approach is as follows:

1. For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.  (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA). 
2. Generate a distance matrix of the nx1 by 2 matrix. Do this for each of the held-out observations. 
3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-|\sqrt{}\. 
5. Apply non-parametric PERMANOVA tests for each holdout observation - this means that on a 100-observation dataset, you apply 100 individual tests where each test uses an independent variable that is equal to 1 for the index of the holdout observation and 0 elsewhere. The hypotheses are specified as such:

+ $H_0$: No differences in map with observation j removed vs. those without j removed 
+ $H_a$: Some difference in map with observation j removed

We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Benjamini-Hochberg False Discovery Rate or Bonferroni corrections (or something similar). We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth adonis test would be small.

# Assessment on Data

## Token Example

We first step through an example of the method as utilized on a relatively small dataset, to show how the algorithm works. Let's consider an example with 63 observations simulated from a multivariate normal distribution.

The code below simulates the data with some clearly abberant values.

```{r simulate_token_data, include = TRUE, message = FALSE, warning = FALSE}
set.seed(34536)
g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
  t() %>%
  data.frame()
g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(25, 20)) %>%
  t() %>%
  data.frame()
g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(25, 30)) %>%
  t() %>%
  data.frame()

yes_structure <- rbind(g1, g2, g3) %>%
  data.frame() %>%
  mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))

yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()

## adds outliers in the last 1 index (simulating from normal based on 5 sd's from mean)
add_outliers <- rbind(rnorm(10, 5, .5)) %>% as_tibble()
names(add_outliers) <- names(yes_structure_scale)

yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
```

1. For each observation in the dataset, remove it, and generate a (t-SNE or MDS) map in 2 dimensions from the resulting (n-1) observations.  (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA). The code below shows this as it pertains to the "multipermanova" function that I have written - it makes use of a function called "insert" from R.utils that adds NAs for each index that is held out.

```{r, eval = TRUE, message = FALSE}
#initializes some stuff 
data = yes_structure_final
n <- nrow(data)
  maplist <- list()
  cmdlist <- list()

#generate maps on holdout data
  for(j in 1:n){

    temp <- Generate_Holdout_Maps(data, j)
    cmdlist[[j]] <- temp$mds
    maplist[[j]] <- temp$tsne

  }

# Insert NAs into the holdout indices 
  maplist_new <- list()
  cmdlist_new <- list()

#replaces holdouts with NAs
for(j in 1:length(maplist)){
    #create temporary vectors with NAs for each of the XY coordingates
    tempX <- R.utils::insert(maplist[[j]][,1], j, NA)
    tempY <- R.utils::insert(maplist[[j]][,2], j, NA)

    tempcmdX <- R.utils::insert(cmdlist[[j]][,1], j, NA)
    tempcmdY <- R.utils::insert(cmdlist[[j]][,2], j, NA)

    maplist_new[[j]] <- tibble(X = tempX, Y = tempY)
    cmdlist_new[[j]] <- tibble(X = tempcmdX, Y = tempcmdY)
}

#prints out the maps
maplist_new[[1]] %>% head()

plot(maplist_new[[1]], pch = 20, main = "Example of maplist (t-SNE), Observation 1 Held Out")

```

2. Generate a distance matrix of the n by 2 matrix. Do this for each of the held-out observations.

3. Vectorize the distance matrix and generate a correlation matrix from the combination of these vectorized distances.

We vectorize distance matrices in the following way, where $d12$ refers to the distance between the first and second observations (in the map). I'm ignoring the upper/lower triangle symmetry in this example. As an example, here's what this looks like in a very simplistic 3 observation case.

$$\mathbf{Dist(X_{nxn})} = \left[\begin{array}
{rrr}
NA & NA & NA \\
NA & 0 & d_{23} \\
NA & d_{32} & 0 
\end{array}\right]
$$

The vectorized form (shown as a row vector) would be: 
$$\mathbf{c(Dist(X))^T} = \left[\begin{array}
{r}
NA & NA & NA &
NA & 0 & d_{32} &
NA & d_{23} & 0
\end{array}\right]
$$
Then, the correlation matrix as follows, where each column comes from the ith holdout map. Note that the below notation reads: $d_{23, 1}$ is the distance between observations 2 and 3 on the map that holds out the 1st observation.

$$ \mathbf{Cor(Dist(X))} = \left[\begin{array}
{rrr}
NA & d_{12, 2} & d_{13,3} \\
NA & NA & d_{23,3} \\
NA & d_{31,2} & NA \\
NA & NA & ...\\
0 & NA & .\\
d_{32,1} & NA & .\\
NA & d_{13,2} & .\\
d_{23,1} & NA & .\\
0 & 0 & .\\
\end{array}\right]
$$

Because of the holdout NAs, we do not have complete cases to compare so instead we use pairwise-complete observations. We can see a visual representation of the NAs for a sample of the distance matrices below:

```{r}
#vectorizes the distances and outputs a list of vectorized distance matrices 
  distmat_list <- lapply(maplist_new, dist)
  distmat_list_cmd <- lapply(cmdlist_new, dist)

levelplot(as.matrix(distmat_list[[1]]), main = paste0("NA Pattern for ",1, "st Holdout"))
levelplot(as.matrix(distmat_list[[3]]), main = paste0("NA Pattern for ",3, "rd Holdout"))
levelplot(as.matrix(distmat_list[[5]]), main = paste0("NA Pattern for ",5, "th Holdout"))

```

The code chunk below shows both steps simultaneously. Note that generating the pearson correlation matrix from the vectorized distance matrices is done using pairwise complete observations - the NAs we include from the holdout set can cause potential problems when this is not specified.

```{r, eval = TRUE}
 #vectorizes the distances and outputs a list of vectorized distance matrices 
  dist_list <- lapply(maplist_new, vectorize_dist)
  dist_list_cmd <- lapply(cmdlist_new, vectorize_dist)

  ## takes the vectorized distance matrices and smashes them into a single data frame - to build correlation matrix

  dist_mat <- dist_list %>% as_tibble(.name_repair = "minimal")
  dist_mat_cmd <- dist_list_cmd %>% as_tibble(.name_repair = "minimal")

  #prints out the data frame of vectorized distance matrices to show what we're looking at
  head(dist_mat)

  #dimensions are n*(n-1)/2 since the distance matrix is triangular - see below
  dim(dist_mat)

```

We calculate the correlation matrix from the above:

```{r}

  #Generates the correlation matrix based on pairwise complete observations, and pearson correlation

  #tsne
  cormat <- cor(dist_mat, use = "pairwise.complete.obs", method = "pearson")
  cormat %>% dim()

  #cmd
  cormat_cmd <- cor(dist_mat_cmd, use = "pairwise.complete.obs", method = "pearson")
  cormat_cmd %>% dim()

```

4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using the result that $D_{ij} = \sqrt{2(1-cor)}$. Note that here, we expect that all the correlation values are greater than 0 - and as shown, they are.

*Note also that with this new distance matrix, we can actually create 2D visualizations (maps) from the distance matrix, creating a "map-of-maps" to show which of the ordinations are different from the rest. (Similarly, you could cluster maps or perform any distance-based method on this matrix to explore differences in the holdout ordinations).*

```{r}
## sanity check - these are all greater than 0
cormat %>% min()
##generate distance on correlation matrix
  cordist <- sqrt(2*(1-cormat)) %>% as.dist()
  cordist_cmd <- sqrt(2*(1-cormat_cmd)) %>% as.dist()
```

5. Apply non-parametric PERMANOVA tests for each holdout observation, using an indicator that is 1 for the holdout observation and 0 for all others.  Cycle the indicator for each index - so you will apply $n$ individual tests.

In the code below, the data does not change - the calls to 'adonis' change only in that the index of the "1" for the indicator variable chanegs at each iteration. Thus, the indicator for the first holdout observation is [1,0,0,...] and so on and so forth.

Note that for adonis, this version uses 1000 permutations rather than a permutation matrix, meaning that this version re-samples the $n+1$ different unique permutations of the indicator which has a single $1$ and $0$'s in all the other indices. *We are working on being a bit more efficient with this permutation strategy so that we do not sample more than necessary*.

```{r}
  nperms = 1000
  model_list <- list()
  model_list_cmd <- list()

 for(j in 1: nrow(as.matrix(cordist))){
    #generate the indicator variable for each feature
    indicator_df <- rep(0, nrow(maplist_new[[1]])) %>% as.data.frame()
    indicator_df_cmd <- rep(0, nrow(cmdlist_new[[1]])) %>% as.data.frame()

    names(indicator_df) <- names(indicator_df_cmd) <- "Indicator"
    indicator_df$Indicator[j] = 1
    indicator_df_cmd$Indicator[j] = 1

     ## create the permutation matrix
  #initialize
permat1 <- matrix(0, nrow = length(indicator_df$Indicator), ncol = length(indicator_df$Indicator))
permat1[1,] <- 1:length(indicator_df$Indicator)

for(i in 2:nrow(permat1)){
  permat1[i,] <- c(i:length(indicator_df$Indicator), 1:(i-1))
}

    #each item in the list is an Adonis test using the indicator for the holdout variable
    model_list[[j]] <- adonis(cordist ~ Indicator, data = indicator_df, permutations = permat1, parallel = 2)
    model_list_cmd[[j]] <- adonis(cordist_cmd ~ Indicator, data = indicator_df_cmd, permutations =  permat1)

  }

```

Note that in the Adonis function, we do obtain a few negative f-statistic values in certain cases. These are the null values based on the permutations used. This seems to occur only for the MDS data - in the cases where we do see negative values, they are quite close to 0.

```{r}

#from t-SNE
model_list[[1]]

model_list[[1]]$f.perms %>% min()
model_list[[1]]$f.perms %>% hist(main = "Example of Permutation Distribution: t-SNE")

#from cmd - an example with negatives included (note that they are all the same value )
model_list_cmd[[1]]$f.perms %>% min()
model_list_cmd[[1]]$f.perms[which(model_list_cmd[[1]]$f.perms<0)]
model_list_cmd[[1]]$f.perms %>% hist(main = "Example of Permutation Distribution: MDS")

```

# Appendix

## Outliers and Influential Points

Similarly, the resulting maps produced by ordination methods like MDS, t-SNE and the like can be impacted by the inclusion or exclusion of points. In short, an "influential" point in a lower dimensional mapping can be defined as one that, when excluded, drives meaningful differences in the ordination method that lead to a different shaped map as compared to when that point is included.

Difference between outliers and influential points.

**talk about influence functions** (Joliffe and cited papers)

Joliffe (2002) overviews tools for identifying influential observations in Pricipal Component Analysis, which is a related dimension-reducting technique to MDS (at least in a Euclidean setting).  However, the literature on more recently-developed methods lacks a similar metric, most notably one that can be used or compared across different mapping methods.

This section overviews some work into identifying multivariate outliers that we've looked into.  We have not found much in the way of users applying Mantel tests or Permanova-type tests on t-SNE or Classical MDS results in the way that we are thinking about things.  In fact, much of the literature around robustness and outliers centers on dealing with the problem by modifying the dimension reduction technique rather than identifying it.

Blouvshtein and Cohen-Or (2018) analyzed the effect of outliers on the maps produced by classical multidimensional scaling methods.  In general, these methods are not particularly robust to the effects of multivariate outlier.  They propose a method to detect outliers prior to the generation of the MDS map so that they can be removed from the data. Their method involves treating the distances input to the algorithm as edges of a complete graph that connects each data point.

These edges $d1, d2, d3$ can then be used to form triangles - in general, outliers tend to *break* the triangle inequality and inliers tend to satisfy it. They recommend a rule to examine the histogram H of $b$, the number of edges of broken triangles, and generate a threshold $\phi$ that is intended to identify inliers vs. outliers.

Two other papers, including Forrero and Giannakis (2012) as well as Kong et al. (2019) frame the problem in a slightly different way. They consider a penalized *stress* metric for optimization, utilizing regularization to better identify outlier points.  In both cases, they are able to identify outliers using a majorization-minimization procedure that iteratively produce more outlier-robust maps of the lower-dimensional embedding of interest.

### 2021-08-21 OutlierComparison.Rmd
Adding files from R

---
title: "Outlier Detection in High-Dimensional Methods"
author: "Paul Harmon"
date: "3/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(janitor)
library(magrittr)
library(knitr)
library(Rtsne)
library(robustHD)
library(GGally)
library(stringr)
library(plotly)
library(R.utils)
library(vegan)
library(mnormt)

```

# Introduction

This document outlines a method for identification of multivariate outliers using several different methods for mapping high-dimensional data in lower-dimensional spaces.  The goal of this analysis is to identify a situation where the results of these maps are distorted or altered by the presence of a multivariate outlier.

## Methods Analyzed

This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE. In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.

Our Permanova-based approach is as follows:

1. For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.  (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA). 
2. Generate a distance matrix of the nx1 by 2 matrix. Do this for each of the held-out observations. 
3. Vectorize the distance matrix and generate a correlation matrix from the combination of these vectorized distances.
4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix. 
5. Apply non-parametric PERMANOVA tests for each holdout observation - this means that on a 100-observation dataset, you apply 100 individual tests where each test uses an independent variable that is equal to 1 for the index of the holdout observation and 0 elsewhere.

We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Bonferroni corrections (or something relatively conservative). We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth adonis test would be small.

# Literature Review

This section overviews some work into identifying multivariate outliers that we've looked into.  We have not found much in the way of users applying Mantel tests or Permanova-type tests on t-SNE or Classical MDS results in the way that we are thinking about things.  In fact, much of the literature around robustness and outliers centers on dealing with the problem rather than identifying it.

Blouvshtein and Cohen-Or (2018) analyzed the effect of outliers on the maps produced by classical multidimensional scaling methods.  In general, these methods are not particularly robust to the effects of multivariate outlier.  They propose a method to detect outliers prior to the generation of the MDS map so that they can be removed from the data. Their method involves treating the distances input to the algorithm as edges of a complete graph that connects each data point.

These edges $d1, d2, d3$ can then be used to form triangles - in general, outliers tend to *break* the triangle inequality and inliers tend to satisfy it. They recommend a rule to examine the histogram H of $b$, the number of edges of broken triangles, and generate a threshold $\phi$ that is intended to identify inliers vs. outliers.

Two other papers, including Forrero and Giannakis (2012) as well as Kong et al. (2019) frame the problem in a slightly different way. They consider a penalized *stress* metric for optimization, utilizing regularization to better identify outlier points.  In both cases, they are able to identify outliers using a majorization-minimization procedure that iteratively produce more outlier-robust maps of the lower-dimensional embedding of interest.

# Data

The data I'm using refers to statistics (att, commpletions, yds, yards per attempt, TD, INT, longest play, sacks, and game points) for NFL quarterbacks in 2016. Since the data are at the game level, I took a subset of only 100 game/QB combinations (to reduce the computation time since this is a proof of concept). Because we are utilizing robust methods for standardizing the data, we remove the INT from consideration.

```{r nfl_readin, echo = TRUE}

nfl <- read_csv("C:/Users/paulh/Downloads/QBStats_2016.csv")
nfl_f <- filter(nfl, att > 5)

nfl_s <- nfl_f %>% dplyr::select(qb, att, cmp, yds, ypa, td, int, lg, sack, game_points)

set.seed(2521)
nfl_subset <- nfl_s[sample(1:nrow(nfl_s), 100),]
```

The code chunk below is set to be ignored. This adds an outlier value, then scales it. We are going to move forward with the opposite approach. 
```{r add_outlier_first, eval = TRUE}
#### Data Scaling: #####
nfl_subset[101,] <- list("Outlier OutlierO. Outlier",95,14,750,2,7,3,99,2,54)
## Scales the data with robust scaling passing centerfun = median and mean abs dev as scale fxn
library(robustHD) #standardize function
nfl_fin2 <- nfl_fin <- nfl_subset %>% select(-c(qb, int)) %>% lapply(standardize, centerFun = median, scaleFun = mad) %>% as_tibble()
```

```{r, scale_then_outlier, eval = FALSE, echo = FALSE}
## Alternatively, scales the outlier and adds the new data
nfl_fin2 <- nfl_fin <- nfl_subset %>% select(-c(qb,int)) %>% lapply(scale) %>% as_tibble()

## adds the outlier on the scaled version
nfl_fin2[101,] <- list(6, 5,6,5,5,6,5,5)

#####
```

```{r nflparcord}
#extracts a readable name for the Name Column
nfl_subset$qbname <- lapply(str_split(nfl_subset$qb, " "),`[[`, 1)  %>% unlist()

## Show a PCP of the players with outlier (on unscaled data)
ggparcoord(as.matrix(nfl_subset), columns = 2:10, group = 11, scale = 'uniminmax') + ggtitle("QB Data: Unscaled") + theme_bw()

nfl_fin2$Group <- c(rep("O",100), "Outlier") %>% factor()
nfl_fin2$Alpha <- ifelse(nfl_fin2$Group == "Outlier", .99, 0.5)

nfl_fin2$Alpha <- ifelse(nfl_fin2$Group == "Outlier", .99, 0.5)

ggparcoord(as.matrix(nfl_fin2), columns = 1:8, groupColumn = 9, alphaLines = 0.8, showPoints = TRUE) + ggtitle("QB Data: Scaled") + theme_bw() + scale_color_viridis_d(direction = -1, end = 0.95)

ggparcoord(as.matrix(nfl_fin2), columns = 1:8, groupColumn = 9,alphaLines = 0.8, showPoints = TRUE, scale = 'uniminmax') + ggtitle("QB Data: Scaled With UniMinMax") + theme_bw() + scale_color_viridis_d(direction = -1, end = 0.95)

```

# Method

The approach is as follows, applied for each method:

+ Hold out each observation and produce a map. 
+ For each map, generate a distance matrix and vectorize it. 
+ Run Permanova test for each of the holdouts
+ Assess p-values to identify the outlier point

# Results

## Classical MDS

The results for the classical MDS method are shown below.

```{r nfl_1, include = TRUE}

n <- nrow(nfl_fin)
maplist <- list()
cmdlist <- list()

#generate t-SNE
for(j in 1:n){

  #hold out a single row of the data (in the jth row)
  #note - have to remove, as t-SNE doesn't like NAs
  nfl_holdout <- nfl_fin[-j,]

  #calculate distances FIRST
  nfl_dist <- dist(nfl_holdout)

  #generate the classical MDS
  mds1 <- cmdscale(nfl_dist, k = 2)
  cmdlist[[j]] <- mds1

  #generate the t-SNE map and store x,y in map list
  x <- Rtsne(nfl_dist, perplexity = 10, pca = FALSE, is_distance = TRUE)
  maplist[[j]] <- x$Y

}

# Maplist is a list of t-SNE results where the j-th index needs to be inserted as an NA
# Ditto for cmdlist except this results from the classical multidimensional scaling

#library(R.utils) #for insert function
maplist_new <- list()
cmdlist_new <- list()

for(j in 1:length(maplist)){
  #create temporary vectors with NAs for each of the XY coordingates
  tempX <- insert(maplist[[j]][,1], j, NA)
  tempY <- insert(maplist[[j]][,2], j, NA)

  tempcmdX <- insert(cmdlist[[j]][,1], j, NA)
  tempcmdY <- insert(cmdlist[[j]][,2], j, NA)

  maplist_new[[j]] <- tibble(X = tempX, Y = tempY)
  cmdlist_new[[j]] <- tibble(X = tempcmdX, Y = tempcmdY)
}

head(maplist_new[[1]])
head(cmdlist_new[[1]])

head(maplist_new[[3]])
head(cmdlist_new[[3]])
```

Now we step through the distance matrix calculation, vectorization, and permanova assessment for both t-SNE and the Classical Multidimensional Scaling.

```{r nfl_step2}
vectorize_dist <- function(df) {
  tmp <- dist(df)
  vec <- c(tmp)
return(vec)}

#vectorizes the distances and outputs a list of vectorized distance matrices 
dist_list <- lapply(maplist_new, vectorize_dist)
dist_list_cmd <- lapply(cmdlist_new, vectorize_dist)

## takes the vectorized distance matrices and smashes them into a single data frame - to build correlation matrix

dist_mat <- dist_list %>% as_tibble(.name_repair = "minimal")
dist_mat_cmd <- dist_list_cmd %>% as_tibble(.name_repair = "minimal")

#had to remove the NAs here to get this to run (as it removes when I try to use complete.obs later on)
#dist_df_na <- sapply(dist_mat, na.omit) %>% as_tibble(.name_repair = "minimal")
#since I'm using pairwise complete obs, I shouldn't need to do this

cormat <- cor(dist_mat, use = "pairwise.complete.obs", method = "pearson")
cormat %>% dim()

cormat_cmd <- cor(dist_mat_cmd, use = "pairwise.complete.obs", method = "pearson")
cormat_cmd %>% dim()

##generate distance on correlation matrix
cordist <- sqrt(2*(1-cormat)) %>% as.dist()
cordist_cmd <- sqrt(2*(1-cormat_cmd)) %>% as.dist()

```

Now we can step through the permanova tests for both the Classical MDS and the t-SNE:

```{r nfl_step3}
#distob <- dist(maplist[[j]])
model_list <- list()
model_list_cmd <- list()

#Step through the permanova
for(j in 1: nrow(as.matrix(cordist))){
  #generate the indicator variable for each feature
  indicator_df <- rep(0, nrow(maplist_new[[1]])) %>% as.data.frame()
  indicator_df_cmd <- rep(0, nrow(cmdlist_new[[1]])) %>% as.data.frame()

  names(indicator_df) <- names(indicator_df_cmd) <- "Indicator"
  indicator_df$Indicator[j] = 1
  indicator_df_cmd$Indicator[j] = 1

  #each item in the list is an Adonis test using the indicator for the holdout variable
  model_list[[j]] <- adonis(cordist ~ Indicator, data = indicator_df)
  model_list_cmd[[j]] <- adonis(cordist_cmd ~ Indicator, data = indicator_df_cmd)
}

pullPval <- function(x){temp <- x$aov.tab$`Pr(>F`[1]; return(temp)}
```

Visualizes the results from the CMDS.
```{r cmds1}
df3 = tibble(Holdout = 1:101, Pvalues = sapply(model_list_cmd, pullPval))
df3$Colors <- ifelse(df3$Pvalues < 0.05, "tomato2", "blue4")
ggplot(df3, aes(Holdout, Pvalues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")

which(df3$Pvalues < 0.05)

```

Visualizes the results from the t-SNE. 
```{r tsne1}
df2 = tibble(Holdout = 1:101, PValues = sapply(model_list, pullPval))
df2$Colors <- ifelse(df2$PValues < 0.05, "tomato2", "blue4")
ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")

which(df2$PValues < 0.05)
```

Interestingly, the Permanova results show significant results for the outlier when examining the CMDS method. For t-SNE, this is not the case, even at different levels of perplexity.

```{r pcplookup}
#look back at a PCP that shows the "significant" multivariate responses 
nfl_fin2$tsnePvals <- df2$Colors
nfl_fin2$cmdPvals <- df3$Colors
ggparcoord(as.matrix(nfl_fin2), columns = 1:8, groupColumn = 11,alphaLines = 0.9, showPoints = TRUE, scale = 'uniminmax') + ggtitle("QB Data: Colored By Significant Permanova Results") + theme_bw() + scale_color_viridis_d(direction = -1, end = 0.95)
```

## A Look at the Maps

Interestingly, the map shows the multivariate outlier quite clearly despite not showing it in the permanova setting.

```{r cmd_outlier}
#plot the CMD with the outlier

cmd1 <- cmdscale(dist(nfl_fin), 2) %>% as_tibble(.name_repair = "unique")
names(cmd1) <- c("X","Y")
cmd1$Index <- 1:nrow(cmd1)

ggplot(cmd1, aes(X,Y)) + geom_label(aes(label = Index)) + theme_bw() + ggtitle("Outlier")

```

The t-SNE, on the other hand, doesn't do show the outlier as clearly, even when run at different levels of perplexity (I tried 5, 10, and 20). The map starts to look more like a uniform spread of points as you get higher on the perplexity scale.

```{r tsne_outlier}
rt1 <- Rtsne(dist(nfl_fin), pca = FALSE, is_distance = TRUE, perplexity = 5)

df1 <- tibble("X" = rt1$Y[,1], "Y" = rt1$Y[,2])
df1$Index = 1:nrow(df1)

ggplot(df1, aes(X,Y, label = Index)) + geom_label() + ggtitle("T-SNE: Perplexity = 5")

```

# Simulated Datasets

Starting with a simulated dataset, we assess the effect of influential points on the MDS and t-SNE maps. The data are simulated using multivariate normal distributions with 3 distinct groups, and then 3 outliers that are pushed, on average, 5 standard deviations from the mean for each of the 10 simulated dimensions.

A quick plot of the simulated data is given below:

```{r simulations}
set.seed(34536)
g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
  t() %>%
  data.frame()
g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(25, 20)) %>%
  t() %>%
  data.frame()
g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(25, 30)) %>%
  t() %>%
  data.frame()

yes_structure <- rbind(g1, g2, g3) %>%
  data.frame() %>%
  mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))

yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()

## adds outliers in the last 3 indices (simulating from normal based on 5 sd's from mean)
add_outliers <- rbind(rnorm(10, 5, .5), rnorm(10, 5, .5), rnorm(10, 5, .5)) %>% as_tibble()
names(add_outliers) <- names(yes_structure_scale)

yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))

yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In")

```

Similar to the NFL dataset, we step through the Permanova process in the chunk below:

```{r sourceoutlierfuctions}
source("OutlierCompFunctions.R")
x3 <- MultiPermanova(yes_structure_final[,1:10])

```

The t-SNE maps do not appear to be affected by the outlier observations.

```{r tsne_sims, include = TRUE}
## test:

df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval))
df2$Colors <- ifelse(df2$PValues < 0.05, "tomato2", "blue4")
ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")

which(df2$PValues < 0.05)
```

On the other hand, the MDS maps are quite sensitive to the outlying three observations, all of which have relatively small p-values (before doing any multiple-testing corrections).

```{r mds_sims, include = TRUE }
df2 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval))
df2$Colors <- ifelse(df2$PValues < 0.05, "tomato2", "blue4")
ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("MDS: P-Values From Adonis")

which(df2$PValues < 0.05)
```

We can ramp up the complexity of the simulations in several ways:

+ Increased/Decreased sample size (note that larger sample sizes slow this process down)
+ More/less  variability in groups
+ More/fewer groups
+ Increase/Decrease the size of the outliers

Note that in **all** simulations I've tried as of now, the t-SNE results have not so much as yielded a spurious small p-value for the t-SNE maps.

# Penguins

The following example goes through the same method with the penguins dataset from the Palmer Penguins dataset (Horst et al, 2000). Size measurements, for various parts of penguins are considered, although this analysis focuses specifically on four: Bill length, Bill depth, Flipper length, and body mass (grams).

We step through the permanova process and also ramp up the number of iterations to 5000 per permutation to generate a more granular set of p-values. Some changes to the steps that we took are listed below:

+ Standardize first, then create an outlier (in particular, this is done to make t-SNE more sensitive)
+ Try a "Godzilla-sized" Penguin that is more than 5 or 6 standard deviations away if a relatively large outlier doesn't work

Note that this dataset is larger than the previous one and therefore takes longer to compile.

```{r, include = TRUE, eval = TRUE}
library(palmerpenguins)
data(penguins) #Loads both penguins and penguins_raw
library(tidyverse)
library(janitor)
penguins_raw <- penguins_raw %>%
      clean_names() #Another way to clean up variable names with spaces etc. from janitor

penguins <- penguins_raw %>%
      dplyr::rename (
         bill_length = culmen_length_mm,
         bill_depth = culmen_depth_mm,
         flipper_length = flipper_length_mm
         ) %>%
      mutate (
         species = word (species, 1), #Just first word
         study_name = factor(study_name),
         species = factor(species),
         individual_id = factor(individual_id)
         )

penguinsX <- penguins %>% mutate(NameID = interaction(species,island,sex, sample_number)) %>% 
   dplyr::select(bill_length, bill_depth, flipper_length, body_mass_g, delta_15_n_o_oo, delta_13_c_o_oo, NameID, species)

penguinsX2 <- penguinsX %>% drop_na()
#rownames(penguinsX2) <- penguinsX2$NameID
```

## Visualizing Penguin Data

We take the approach of first selecting four features from the penguin dataset: Bill length, bill depth, flipper length, and body mass.

```{r, eval = TRUE}

penguins_s <- penguinsX2 %>% select(bill_length, bill_depth, flipper_length, body_mass_g)

set.seed(2521)

## Scales the data with robust scaling passing centerfun = median and mean abs dev as scale fxn
library(robustHD) #standardize function
#penguin2 <-  penguins_s %>% lapply(standardize, centerFun = median, scaleFun = mad) %>% as_tibble()

## we actually don't need the robust standardization function here, will add outlier after scaling
penguin2 <- penguins_s %>% lapply(scale) %>% as_tibble()

## Adds outlier Penguin
penguin2[nrow(penguin2) + 1,] <- list(5,6,5,6) # should add values on a 5 or 6 standard deviation scale away
```

```{r, eval = FALSE}
## Optional: Godzilla Penguin
penguin2[nrow(penguin2) + 1, ] <- list(100,99,100,99) #ginormous penguin

```

```{r plotpenguins, eval = TRUE}
#dataset for plot
penguin2p <- penguin2
penguin2p$Species <- c(penguinsX2$species,"Outlier")

## globalminmax shouldn't do any scaling but it does something weird here
ggparcoord(as.matrix(penguin2p), columns = 1:4, group = 5, scale = 'globalminmax') + ggtitle("Pengunin Data") + theme_bw()

## 
penguin2p$ID <- 1:nrow(penguin2p)
penguin2p %>% pivot_longer(1:4) %>% ggplot(aes(name, value, group = ID,color = Species)) + geom_line() + ggtitle("Penguin Data with Outlier Added In")

```

The code chunk below generates the t-SNE maps and classical MDS. In the t-SNE here, we stick with the original number of iterations.

```{r, eval = TRUE}
data = penguin2 
maplist <- list()
  cmdlist <- list()
perp = 20

 #generate maps on holdout data
  for(j in 1:nrow(data)){

    #hold out a single row of the data (in the jth row)
    #note - have to remove, as t-SNE doesn't like NAs
    data_holdout <- data[-j,]

    #calculate distances FIRST
    data_dist <- dist(data_holdout)

    #generate the classical MDS
    mds1 <- cmdscale(data_dist, k = 2)
    cmdlist[[j]] <- mds1

    #generate the t-SNE map and store x,y in map list
    x <- Rtsne(data_dist, perplexity = perp, pca = FALSE, is_distance = TRUE)
    maplist[[j]] <- x$Y
  }
```

Code chunk below inserts NAs into each holdout index, as above. Then vectorizes distances and builds a single dataframe/tibble for each.

```{r, eval = TRUE}
 # Insert NAs into the holdout indices
  maplist_new <- list()
  cmdlist_new <- list()

  for(j in 1:length(maplist)){
    #create temporary vectors with NAs for each of the XY coordinates
    tempX <- insert(maplist[[j]][,1], j, NA)
    tempY <- insert(maplist[[j]][,2], j, NA)

    tempcmdX <- insert(cmdlist[[j]][,1], j, NA)
    tempcmdY <- insert(cmdlist[[j]][,2], j, NA)

    maplist_new[[j]] <- tibble(X = tempX, Y = tempY)
    cmdlist_new[[j]] <- tibble(X = tempcmdX, Y = tempcmdY)
  }

  #vectorizes the distances and outputs a list of vectorized distance matrices 
  dist_list <- lapply(maplist_new, vectorize_dist)
  dist_list_cmd <- lapply(cmdlist_new, vectorize_dist)

  ## takes the vectorized distance matrices and smashes them into a single data frame - to build correlation matrix

  dist_mat <- dist_list %>% as_tibble(.name_repair = "minimal")
  dist_mat_cmd <- dist_list_cmd %>% as_tibble(.name_repair = "minimal")
```

Next steps: Generates the correlation matrices using pairwise complete observations. Then generates distance matrices from them.

```{r, eval = TRUE, cache = TRUE}
 cormat <- cor(dist_mat, use = "pairwise.complete.obs", method = "pearson")
  cormat %>% dim()

  cormat_cmd <- cor(dist_mat_cmd, use = "pairwise.complete.obs", method = "pearson")
  cormat_cmd %>% dim()

  ##generate distance on correlation matrix
  cordist <- sqrt(2*(1-cormat)) %>% as.dist()
  cordist_cmd <- sqrt(2*(1-cormat_cmd)) %>% as.dist()

  model_list <- list()
  model_list_cmd <- list()
```

Finally, we run through the Adonis piece to get each of the individual tests. A 'significant' p-value should indicate evidence that the map is suitably different from the other maps, meaning that if we have a really large outlier, we should be able to identify it.

```{r, eval = TRUE, cache = TRUE}

nperms = 10000
 #Step through the permanova
for(j in 1: nrow(as.matrix(cordist))){
  #generate the indicator variable for each feature
  indicator_df <- rep(0, nrow(maplist_new[[1]])) %>% as.data.frame()
  indicator_df_cmd <- rep(0, nrow(cmdlist_new[[1]])) %>% as.data.frame()

  names(indicator_df) <- names(indicator_df_cmd) <- "Indicator"
  indicator_df$Indicator[j] = 1
  indicator_df_cmd$Indicator[j] = 1

  #each item in the list is an Adonis test using the indicator for the holdout variable
  model_list[[j]] <- adonis(cordist ~ Indicator, data = indicator_df, permutations = nperms)
  model_list_cmd[[j]] <- adonis(cordist_cmd ~ Indicator, data = indicator_df_cmd, permutations = nperms)
}

```

Visualize TSNE:

```{r, eval = TRUE}
df2 = tibble(Holdout = 1:length(model_list), PValues = sapply(model_list, pullPval))
df2$Colors <- ifelse(df2$PValues < 0.05, "tomato2", "blue4")
ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")

which(df2$PValues < 0.05)
```

Visualize CMDS: 
```{r, eval = TRUE}
df3 = tibble(Holdout = 1:length(model_list_cmd), Pvalues = sapply(model_list_cmd, pullPval))
df3$Colors <- ifelse(df3$Pvalues < 0.05, "Significant", "Not-Significant")
ggplot(df3, aes(Holdout, Pvalues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")

which(df3$Pvalues < 0.05)

```

Visualize the PCP
```{r, eval = TRUE}

x <- penguin2p %>% mutate(Pvalues = df3$Pvalues, PvalColors = df3$Colors) %>% pivot_longer(1:4)

#generate plot
ggplot(x, aes(name, value, group = ID,shape = factor(Species), color = factor(PvalColors))) + geom_line() + ggtitle("Penguin Data with Outlier Added In")

#look at tabular view
table(x$Species, x$PvalColors)

```

### 2021-08-21 Permanova Real Data Section.Rmd
Adding files from R

---
title: 'Permanova: Real Data Analysis'
author: "Paul Harmon"
date: "7/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("OutlierCompFunctions.R")
library(ggplot2)
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(janitor)
library(magrittr)
library(knitr)
library(Rtsne)
library(robustHD)
library(GGally)
library(stringr)
library(plotly)
library(R.utils)
library(vegan)
library(mnormt)
```

# Introduction

This document overviews the real data analysis portion of the t-SNE/MDS/Permanova paper. We are currently looking at two specific datasets: Palmer Penguins and Carnegie Classifications.

# Carnegie

Looks at the 2015 Carnegie Classification data - without doing much, to see how the data are changed.

## Raw Data

```{r, cache = TRUE}
CC2015 <- read.csv("https://raw.githubusercontent.com/paulharmongj/Carnegie_SEM/master/data/CC2015data.csv")

CC2015_filter <- CC2015 %>% filter(BASIC2015 %in% c(15,16,17)) %>% select(PDNFRSTAFF, S.ER.D, NONS.ER.D, FACNUM, HUM_RSD, STEM_RSD, OTHER_RSD, SOCSC_RSD)
perplexity = 30

## NEED TO STANDARDIZE THE DATA FIRST (MAKE UNITLESS)
CCscale <- CC2015_filter %>% lapply(scale) %>% as_tibble()

out <- MultiPermanova(CCscale)

```

```{r}
#generate some plots of the output
df3 = tibble(Holdout = 1:length(out$cmdlist), PValues = sapply(out$cmdlist, pullPval), Fstat = sapply(out$cmdlist, pullFstat))
df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)

  #Visualize the results and store in a list
cc <- CC2015 %>% filter(BASIC2015 %in% c(15,16,17)) 
  cc$NAME[which(df3$PValues < 0.05)]
  df3$Name <- cc$NAME
  sigplot2 <- ggplot(df3, aes(Holdout, Fstat, label = Name)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
  sigplot2 %>% ggplotly()

```

## Scaled with Per-Capita Features

This version looks at the same input variables but considers the set of features that are scaled in the per-capita version. Due to high degrees of correlation between the raw inputs and some of their per-capita counterparts, I only included the per-capita versions for STEM expenditures, Non-STEM research expenditures, and tenurable research faculty count.

```{r, cache = TRUE, eval = FALSE}
CC2015_filter_PC <- CC2015_filter %>% mutate(S.ER.DPC = S.ER.D/FACNUM, NONS.ER.DPC = NONS.ER.D/FACNUM, STAFFPC = PDNFRSTAFF/FACNUM)

out2 <- MultiPermanova(CC2015_filter_PC)

```

## Plots with Permutations

```{r}
library(tidyverse)
library(ggplot2)
library(ggridges)

histlist1 <- list()
histlist2 <- list()

for(j in 1:length(out$tsnelist)){
  histlist1[[j]] <- out$tsnelist[[j]]$f.perms %>% as_tibble() %>% ggplot(aes(V1)) + geom_histogram(bins = 10)
  histlist2[[j]] <-  out$cmdlist[[j]]$f.perms %>% as_tibble() %>% ggplot(aes(V1)) + geom_density() + scale_x_log10()
}

f1Fs_long %>% ggplot(mapping=aes(y=Number, x=value, fill = factor(stat(quantile)))) +
    stat_density_ridges(geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 0.95, 
    alpha=0.3,
    scale=2
    ) +
    scale_fill_manual(
    name = "Probability", values = c("#A0A0A0A0", "#FF0000A0"),
    labels = c("(0, 0.95]", "(0.95, 1]")
    )  +
   scale_x_continuous(limits = c(-1, 3.5)) +
    geom_point(data=observed, mapping=aes(y=Number, x=FA), inherit.aes = F) +
  geom_line(data=observed, mapping=aes(y=Number, x=FA), inherit.aes=F) +
    coord_flip() + 
    theme_ridges() + 
  geom_vline(xintercept=0, col="grey")

```

## "Extreme" points added in

The goal here is to add a couple "extreme" points and see how this method handles the original "influential" points.

```{r, eval = FALSE}

```

# Palmer

### 2021-08-21 Permanova-Real-Data-Section.Rmd
Adding files from R

--- 2021-08-21 Permanova Real Data Section.Rmd
@@
 author: "Paul Harmon"
-date: "7/13/2021"
+date: "6/26/2021"
 output: html_document
@@
 
-## NEED TO STANDARDIZE THE DATA FIRST (MAKE UNITLESS)
-CCscale <- CC2015_filter %>% lapply(scale) %>% as_tibble()
-
-out <- MultiPermanova(CCscale)
-
+out <- MultiPermanova(CC2015_filter)
 ```
@@
   #Visualize the results and store in a list
+  sigplot2 <- ggplot(df3, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
+  sigplot2
+
   cc <- CC2015 %>% filter(BASIC2015 %in% c(15,16,17)) 
   cc$NAME[which(df3$PValues < 0.05)]
-  df3$Name <- cc$NAME
-  sigplot2 <- ggplot(df3, aes(Holdout, Fstat, label = Name)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
-  sigplot2 %>% ggplotly()
 
@@
 
-## Scaled with Per-Capita Features
+## Scaled
 
@@
 
-```{r, cache = TRUE, eval = FALSE}
+```{r, cache = TRUE
+}
 CC2015_filter_PC <- CC2015_filter %>% mutate(S.ER.DPC = S.ER.D/FACNUM, NONS.ER.DPC = NONS.ER.D/FACNUM, STAFFPC = PDNFRSTAFF/FACNUM)
@@
 out2 <- MultiPermanova(CC2015_filter_PC)
-
-```
-
-## Plots with Permutations
-
-```{r}
-library(tidyverse)
-library(ggplot2)
-library(ggridges)
-
-histlist1 <- list()
-histlist2 <- list()
-
-for(j in 1:length(out$tsnelist)){
-  histlist1[[j]] <- out$tsnelist[[j]]$f.perms %>% as_tibble() %>% ggplot(aes(V1)) + geom_histogram(bins = 10)
-  histlist2[[j]] <-  out$cmdlist[[j]]$f.perms %>% as_tibble() %>% ggplot(aes(V1)) + geom_density() + scale_x_log10()
-}
-
-f1Fs_long %>% ggplot(mapping=aes(y=Number, x=value, fill = factor(stat(quantile)))) +
-    stat_density_ridges(geom = "density_ridges_gradient",
-    calc_ecdf = TRUE,
-    quantiles = 0.95, 
-    alpha=0.3,
-    scale=2
-    ) +
-    scale_fill_manual(
-    name = "Probability", values = c("#A0A0A0A0", "#FF0000A0"),
-    labels = c("(0, 0.95]", "(0.95, 1]")
-    )  +
-   scale_x_continuous(limits = c(-1, 3.5)) +
-    geom_point(data=observed, mapping=aes(y=Number, x=FA), inherit.aes = F) +
-  geom_line(data=observed, mapping=aes(y=Number, x=FA), inherit.aes=F) +
-    coord_flip() + 
-    theme_ridges() + 
-  geom_vline(xintercept=0, col="grey")
 

### 2021-08-21 Simulation Study Document II.Rmd
Adding files from R

---
title: "Simulation Study Document"
author: "Paul Harmon"
date: "5/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("OutlierCompFunctions.R")
library(ggplot2)
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(janitor)
library(magrittr)
library(knitr)
library(Rtsne)
library(robustHD)
library(GGally)
library(stringr)
library(plotly)
library(R.utils)
library(vegan)
library(mnormt)

```

# Introduction

This doc contains simulation studies around the efficacy of a Permanova-based method for identifying influential points in a t-SNE or MDS map.  The document steps through a process of simulating data from 3 different groups (currently, these groups are clearly separated) with *3* influential points simulated on a standardized scale 5-6 standard deviations above the observed data.

The code below outlines the number of simulations run - currently this number is small for computational ease but will be increased (with the use of adaptive Permanova methods).

```{r}
n_sims = 100

#create a bunch of lists to store stuff later
plot_data_list <- list()
plot_tsne_results_list <- list()
plot_mds_results_list <- list()
plot_tsne_fstat_results_list <- list()
plot_mds_fstat_results_list <- list()
results_list <- list()
df3list = df2list = list()
```

The below chunk runs the simulations. Each iteration simulates a new (contaminated) dataset, runs the t-SNE and MDS mapping procedure, and tests the consistency of the configurations for each of the *n* holdout points. **This code is not run in this version.**

```{r original_sims, eval = FALSE, cache = TRUE, warning = FALSE, message = FALSE}

for(j in 1:n_sims){

  #### Simulate the Data

  g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(25, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(25, 30)) %>%
    t() %>%
    data.frame()

  yes_structure <- rbind(g1, g2, g3) %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))

  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()

  ## adds outliers in the last 3 indices (simulating from normal based on 5 sd's from mean)
  add_outliers <- rbind(rnorm(10, 5, .5), rnorm(10, 5, .5), rnorm(10, 5, .5)) %>% as_tibble()
  names(add_outliers) <- names(yes_structure_scale)

  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))

  #Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In")
  plot_data_list[[j]] <- datplot

  #### Run the Permanova mapping

  x3 <- MultiPermanova(yes_structure_final[,1:10])

  results_list[[j]] <- x3

  df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval), Fstat = sapply(x3$tsnelist, pullFstat))
  df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
  df2list[[j]] = df2

  #### TSNE Plots########################################################
  #Visualize the results and store in a list
  sigplot <- ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")
  plot_tsne_results_list[[j]] <- sigplot

  #Visualize the f stats and store in a list
  sigplotf <- ggplot(df2, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: F Stats From Adonis")
  plot_tsne_fstat_results_list[[j]] <- sigplotf

  ### CMD Plots #########################################################
  df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval), Fstat = sapply(x3$cmdlist, pullFstat))
    df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)
  df3list[[j]] = df3

  #Visualize the results and store in a list
  sigplot2 <- ggplot(df3, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
  plot_mds_results_list[[j]] <- sigplot2

  #Visualize the f stats and store in a list
  sigplot2f <- ggplot(df3, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: F Stats From Adonis")
  plot_mds_fstat_results_list[[j]] <- sigplot2f

}
```

This chunk runs the simulation with slighlty increased variability in the three groups, and a restructuring of the influential points. Rather than simulating as outliers all on the positive side, I simulate one that has a large positive set of outlying points, one with a large negative set of outlying points, and one that simply 'bounces around' the multivariate space and breaks the grouping structure.

```{r sims_v2, eval = TRUE, cache = TRUE, warning = FALSE, message = FALSE}

for(j in 1:n_sims){

  #### Simulate the Data with outliers adn a single group (simulates all from one group)

  g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(45, 10)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(45, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(45, 30)) %>%
    t() %>%
    data.frame()

  yes_structure <- rbind(g1, g2, g3) %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))

  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()

  ## adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
  add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
  names(add_outliers) <- names(yes_structure_scale)

  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))

  #Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
  plot_data_list[[j]] <- datplot

  #### Run the Permanova mapping

  x3 <- MultiPermanova(yes_structure_final[,1:10])

  results_list[[j]] <- x3

  df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval), Fstat = sapply(x3$tsnelist, pullFstat))
  df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
  df2list[[j]] = df2

  #### TSNE Plots########################################################
  #Visualize the results and store in a list
  sigplot <- ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")
  plot_tsne_results_list[[j]] <- sigplot

  #Visualize the f stats and store in a list
  sigplotf <- ggplot(df2, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: F Stats From Adonis")
  plot_tsne_fstat_results_list[[j]] <- sigplotf

  ### CMD Plots #########################################################
  df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval), Fstat = sapply(x3$cmdlist, pullFstat))
    df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)
  df3list[[j]] = df3

  #Visualize the results and store in a list
  sigplot2 <- ggplot(df3, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
  plot_mds_results_list[[j]] <- sigplot2

  #Visualize the f stats and store in a list
  sigplot2f <- ggplot(df3, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: F Stats From Adonis")
  plot_mds_fstat_results_list[[j]] <- sigplot2f

}
```

# Results {.tabset}

The simulated datasets can be seen in the tabset below.

## PlotList
```{r}
library(ggpubr)
#ggarrange(plotlist = plot_data_list)
print(plot_data_list)
```

## Summary Metrics

We can start to look at some results below. The code generates the rate of correctly classified influential points for both the MDS (shown first) and t-SNE maps. The MDS correctly classifies the influential point *some* of the time, but the t-SNE method very rarely classifies the influential points as having an impact on the configuration.

```{r}

#print the head of the df lists
df3list %>% head(3)
### Identify the number of CORRECTLY Classified influential points
#extract the p-values from the last 3 from MDS
influential_correct <- function(temp){length(which(tail(temp, 3)$PValues < 0.05))/3}
mds_correct <- lapply(df3list, influential_correct)

# extract the pvalues from the last 3 of the t-SNE
tsne_correct <- lapply(df2list, influential_correct)

dat = tibble(MDS = unlist(mds_correct), TSNE = unlist(tsne_correct))
#plot(dat$MDS, dat$TSNE)

mds_correct %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified")
tsne_correct %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified")

```

Now we can look at the rate of misclassified non-influential points. These are the 60 'regular' points that are spuriously detected as outliers. Each percentage is calculated as a percentage of the 60 that are NOT small p-values (currenlty < 0.05 without a correction). We want a high percentage of correctly-identified non-influential points (although given the machinery of the hypothesis tests used, we would expect at least some spurious detections).

Both are relatively high percentages (as we would expect) - but the t-SNE hovers right around the expected 0.95 and MDS is higher.

```{r}

### Identify rate of CORRECTLY Classified non-influential points
#extract the p-values from the last 3 from MDS
ni_correct <- function(temp){length(which(head(temp, 60)$PValues >= 0.05))/60}

correct_mds <- lapply(df3list, ni_correct)
#kable(correct_mds, col.names = "Correctly Identified Non-Influential: MDS")

# extract the pvalues from the last 3 of the t-SNE
correct_tsne <- lapply(df2list, ni_correct)
#kable(correct_tsne, col.names = "Correctly Identified Non-Influential: tSNE")

correct_mds %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified Non-Influential")
correct_tsne %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified Non-Influential")

```

# Plots {.tabset}

## MDS P Values

Plots are shown below for the MDS .

```{r}
plot_mds_results_list
```

## MDS F Stats

```{r}
plot_mds_fstat_results_list
```

## tSNE

```{r}
plot_tsne_results_list
```

## t-SNE F Stats

```{r}
plot_tsne_fstat_results_list
```

### 2021-08-21 Simulation Study Document V3.Rmd
Adding files from R

--- 2021-08-21 Simulation Study Document II.Rmd
@@
 ---
-title: "Simulation Study Document"
+title: "Simulation Study Document V4"
 author: "Paul Harmon"
-date: "5/20/2021"
+date: "`r format(Sys.time(), '%B %d, %Y')`"
 output: html_document
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
 

### 2021-08-21 Simulation Study Document.Rmd
Adding files from R

--- 2021-08-21 Simulation Study Document II.Rmd
@@
 This doc contains simulation studies around the efficacy of a Permanova-based method for identifying influential points in a t-SNE or MDS map.
-The document steps through a process of simulating data from 3 different groups (currently, these groups are clearly separated) with *3* influential points simulated on a standardized scale 5-6 standard deviations above the observed data.
-
-The code below outlines the number of simulations run - currently this number is small for computational ease but will be increased (with the use of adaptive Permanova methods).
 
 ```{r}
-n_sims = 100
+n_sims = 10
+```
 
-#create a bunch of lists to store stuff later
+```{r, cache = TRUE, warning = FALSE, message = FALSE}
 plot_data_list <- list()
@@
 plot_mds_results_list <- list()
-plot_tsne_fstat_results_list <- list()
-plot_mds_fstat_results_list <- list()
 results_list <- list()
-df3list = df2list = list()
-```
-
-The below chunk runs the simulations.
-Each iteration simulates a new (contaminated) dataset, runs the t-SNE and MDS mapping procedure, and tests the consistency of the configurations for each of the *n* holdout points.
-**This code is not run in this version.**
-
-```{r original_sims, eval = FALSE, cache = TRUE, warning = FALSE, message = FALSE}
 
@@
   #### Simulate the Data
-
+  set.seed(34536)
   g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
@@
   df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
-  df2list[[j]] = df2
 
-  #### TSNE Plots########################################################
   #Visualize the results and store in a list
@@
 
-  #Visualize the f stats and store in a list
-  sigplotf <- ggplot(df2, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: F Stats From Adonis")
-  plot_tsne_fstat_results_list[[j]] <- sigplotf
-
-  ### CMD Plots #########################################################
   df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval), Fstat = sapply(x3$cmdlist, pullFstat))
   df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)
-  df3list[[j]] = df3
 
   #Visualize the results and store in a list
-  sigplot2 <- ggplot(df3, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
+  sigplot2 <- ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")
   plot_mds_results_list[[j]] <- sigplot2
-
-  #Visualize the f stats and store in a list
-  sigplot2f <- ggplot(df3, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: F Stats From Adonis")
-  plot_mds_fstat_results_list[[j]] <- sigplot2f
 
@@
 
-This chunk runs the simulation with slighlty increased variability in the three groups, and a restructuring of the influential points.
-Rather than simulating as outliers all on the positive side, I simulate one that has a large positive set of outlying points, one with a large negative set of outlying points, and one that simply 'bounces around' the multivariate space and breaks the grouping structure.
+We can start to look at some results below. The code generates the rate of correctly classified influential points for both the MDS (shown first) and t-SNE maps.
 
-```{r sims_v2, eval = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
-
-for(j in 1:n_sims){
-
-  #### Simulate the Data with outliers adn a single group (simulates all from one group)
-
-  g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(45, 10)) %>%
-    t() %>%
-    data.frame()
-  g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(45, 20)) %>%
-    t() %>%
-    data.frame()
-  g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(45, 30)) %>%
-    t() %>%
-    data.frame()
-
-  yes_structure <- rbind(g1, g2, g3) %>%
-    data.frame() %>%
-    mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))
-
-  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
-
-  ## adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
-  add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
-  names(add_outliers) <- names(yes_structure_scale)
-
-  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
-  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))
-
-  #Visualize the data and store in a list
-  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
-  plot_data_list[[j]] <- datplot
-
-  #### Run the Permanova mapping
-
-  x3 <- MultiPermanova(yes_structure_final[,1:10])
-
-  results_list[[j]] <- x3
-
-  df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval), Fstat = sapply(x3$tsnelist, pullFstat))
-  df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
-  df2list[[j]] = df2
-
-  #### TSNE Plots########################################################
-  #Visualize the results and store in a list
-  sigplot <- ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")
-  plot_tsne_results_list[[j]] <- sigplot
-
-  #Visualize the f stats and store in a list
-  sigplotf <- ggplot(df2, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: F Stats From Adonis")
-  plot_tsne_fstat_results_list[[j]] <- sigplotf
-
-  ### CMD Plots #########################################################
-  df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval), Fstat = sapply(x3$cmdlist, pullFstat))
-    df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)
-  df3list[[j]] = df3
-
-  #Visualize the results and store in a list
-  sigplot2 <- ggplot(df3, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
-  plot_mds_results_list[[j]] <- sigplot2
-
-  #Visualize the f stats and store in a list
-  sigplot2f <- ggplot(df3, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: F Stats From Adonis")
-  plot_mds_fstat_results_list[[j]] <- sigplot2f
-
-}
+```{r}
+### Identify the number of CORRECTLY Classified influential points
+#extract the p-values from the last 3 from MDS
+length(which(tail(df3, 3)$PValues < 0.05))/3
+# extract the pvalues from the last 3 of the t-SNE
+length(which(tail(df2, 3)$Pvalues < 0.05))/3
 ```
 
-# Results {.tabset}
-
-The simulated datasets can be seen in the tabset below.
-
-## PlotList
-```{r}
-library(ggpubr)
-#ggarrange(plotlist = plot_data_list)
-print(plot_data_list)
-```
-
-## Summary Metrics
-
-We can start to look at some results below.
-The code generates the rate of correctly classified influential points for both the MDS (shown first) and t-SNE maps.
-The MDS correctly classifies the influential point *some* of the time, but the t-SNE method very rarely classifies the influential points as having an impact on the configuration.
-
-```{r}
-
-#print the head of the df lists
-df3list %>% head(3)
-### Identify the number of CORRECTLY Classified influential points
-#extract the p-values from the last 3 from MDS
-influential_correct <- function(temp){length(which(tail(temp, 3)$PValues < 0.05))/3}
-mds_correct <- lapply(df3list, influential_correct)
-
-# extract the pvalues from the last 3 of the t-SNE
-tsne_correct <- lapply(df2list, influential_correct)
-
-dat = tibble(MDS = unlist(mds_correct), TSNE = unlist(tsne_correct))
-#plot(dat$MDS, dat$TSNE)
-
-mds_correct %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified")
-tsne_correct %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified")
-
-```
-
-Now we can look at the rate of misclassified non-influential points.
-These are the 60 'regular' points that are spuriously detected as outliers.
-Each percentage is calculated as a percentage of the 60 that are NOT small p-values (currenlty < 0.05 without a correction).
-We want a high percentage of correctly-identified non-influential points (although given the machinery of the hypothesis tests used, we would expect at least some spurious detections).
-
-Both are relatively high percentages (as we would expect) - but the t-SNE hovers right around the expected 0.95 and MDS is higher.
+Now we can look at the rate of misclassified non-influential points. These are the 60 'regular' points that are spuriously detected as outliers.
 
@@
 #extract the p-values from the last 3 from MDS
-ni_correct <- function(temp){length(which(head(temp, 60)$PValues >= 0.05))/60}
-
-correct_mds <- lapply(df3list, ni_correct)
-#kable(correct_mds, col.names = "Correctly Identified Non-Influential: MDS")
-
+length(which(head(df3, 60)$PValues >= 0.05))/60 
 # extract the pvalues from the last 3 of the t-SNE
-correct_tsne <- lapply(df2list, ni_correct)
-#kable(correct_tsne, col.names = "Correctly Identified Non-Influential: tSNE")
-
-correct_mds %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified Non-Influential")
-correct_tsne %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified Non-Influential")
-
+length(which(tail(df2, 60)$PValues >= 0.05))/60
 ```
@@
 
-## MDS P Values
+## MDS
 
@@
 
-## MDS F Stats
-
-```{r}
-plot_mds_fstat_results_list
-```
-
 ## tSNE
@@
 ```
-
-## t-SNE F Stats
-
-```{r}
-plot_tsne_fstat_results_list
-```

### 2021-08-21 Simulation-Study-Document-II.Rmd
Adding files from R

--- 2021-08-21 Simulation Study Document II.Rmd
@@
 Each iteration simulates a new (contaminated) dataset, runs the t-SNE and MDS mapping procedure, and tests the consistency of the configurations for each of the *n* holdout points.
-**This code is not run in this version.**
 
@@
 ```
-
-This chunk runs the simulation with slighlty increased variability in the three groups, and a restructuring of the influential points.
-Rather than simulating as outliers all on the positive side, I simulate one that has a large positive set of outlying points, one with a large negative set of outlying points, and one that simply 'bounces around' the multivariate space and breaks the grouping structure.
 

### 2021-08-21 Simulation-Study-Document.Rmd
Adding files from R

--- 2021-08-21 Simulation Study Document.Rmd
@@
 ```{r}
-n_sims = 10
+n_sims = 100
 ```
@@
 
-  df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval), Fstat = sapply(x3$tsnelist, pullFstat))
+  df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval))
   df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
@@
 
-  df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval), Fstat = sapply(x3$cmdlist, pullFstat))
+  df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval))
   df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)

### 2021-08-21 tSNE Introduction Document.Rmd
Adding files from R

---
title: "TSNE Introduction"
author: "Paul Harmon"
date: "6/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rtsne)
```

# Introduction

This chapter should act as a pseudo-paper that breaks out the derviation and mathematical details of t-SNE in detail. Additionally, should overview the pros and cons of the method and discuss how it should be used.

This will act as an addendum to my introduction section but should also be submittable as a stand-alone paper.

## How is t-SNE used

## Comparison methods

### SNE
### MDS/Sammon Mapping
### Umap

# t-SNE Algorithm and Math

## Derivation

## Mathematical Details

## PCA as Initial Step

## Considerations

# Perplexity

## Cost Decreases as Function of Perplexity

## Perplexity

### 2021-08-21 OutlierCompFunctions.R

@@ -1,0 +1,164 @@
+## Script to functionalize this process
+
+
+# preliminary functions: 
+
+## Vectorize Dist: Function that takes a dataframe and returns a vectorized version of a distance matrix applied to that data matrix. 
+## This function is used for generating vectorized data when we step through the Mantel test. 
+
+vectorize_dist <- function(df) {
+  tmp <- dist(df)
+  vec <- c(tmp)
+  return(vec)}
+
+## Pull P-Val and Pull F stat return the p-values and f-statstics from the list of output MultiPermanova puts out. 
+
+pullPval <- function(x){temp <- x$aov.tab$`Pr(>F`[1]; return(temp)}
+
+pullFstat <- function(x){temp <- x$aov.tab$`F.Model`[1]; return(temp)}
+
+
+
+### Dependent Functions
+
+#generate maps on holdout data (jth iteration is held out)
+# this function should be able to operate on multiple mapping techniques or a single one 
+
+Generate_Holdout_Maps <- function(data, j, ...){  
+  
+  #hold out a single row of the data (in the jth row)
+  #note - have to remove, as t-SNE doesn't like NAs
+  data_holdout <- data[-j,]
+  
+  #calculate distances FIRST
+  data_dist <- dist(data_holdout)
+  
+  #### Creates Maps (based on several methods)
+  
+  #generate the classical MDS
+  mds1 <- cmdscale(data_dist, k = 2)
+  
+  #generate the t-SNE map and store x,y in map list
+  x <- Rtsne(data_dist, perplexity = perplexity, pca = pca_options, is_distance = TRUE)
+  
+  return(list(mds = mds1, tsne = x$Y))
+}
+
+
+
+## Generates the Permanova test 
+# 1: listobject: list of maps (maplist, cmdlist) to evaluate
+# 2: nperms - should inherit from above
+# 3: j: index to be looped through 
+# HDpermanova <- function(listobject, nperms, j){
+#   #generate the indicator variable for each feature
+#   indicator_df <- rep(0, nrow(listobject[[1]])) %>% as.data.frame()
+#   
+#   names(indicator_df) <- "Indicator"
+#   indicator_df$Indicator[j] = 1
+#   
+#   #each item in the list is an Adonis test using the indicator for the holdout variable
[…]

### 2021-08-21 OutlierCompFunctions_TestCases.R

@@ -1,0 +1,79 @@
+## Test Cases Four Outlier Comp Functions
+
+x1 <- MultiPermanova(nfl_fin2)
+x2 <- MultiPermanova(penguinsX2[,1:4])
+
+
+
+
+## Simulated Data (with clear structure and 3 major ourliers)
+
+
+
+# "Clear Structure": with 10, 20, and 30 observations per group
+set.seed(34536)
+g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
+  t() %>%
+  data.frame()
+g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(25, 20)) %>%
+  t() %>%
+  data.frame()
+g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(25, 30)) %>%
+  t() %>%
+  data.frame()
+
+
+yes_structure <- rbind(g1, g2, g3) %>%
+  data.frame() %>%
+  mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))
+
+yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
+
+## adds outliers in the last 3 indices (simulating from normal based on 5 sd's from mean)
+add_outliers <- rbind(rnorm(10, 5, .5), rnorm(10, 5, .5), rnorm(10, 5, .5)) %>% as_tibble()
+names(add_outliers) <- names(yes_structure_scale)
+
+
+yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
+yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))
+
+
+
+x3 <- MultiPermanova(yes_structure_final[,1:10])
+
+
+
+## test: 
+
+df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval), Fstat = sapply(x3$tsnelist, pullFstat))
+df2$Colors <- ifelse(df2$PValues < 0.05, "tomato2", "blue4")
+ggplot(df2, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: F Stats From Adonis")
+
+
+which(df2$PValues < 0.05)
+
[…]

### 2021-08-21 OutlierCompFunctions_old.R

@@ -1,0 +1,126 @@
+## Script to functionalize this process
+
+
+# preliminary functions: 
+
+## Vectorize Dist: Function that takes a dataframe and returns a vectorized version of a distance matrix applied to that data matrix. 
+## This function is used for generating vectorized data when we step through the Mantel test. 
+
+vectorize_dist <- function(df) {
+  tmp <- dist(df)
+  vec <- c(tmp)
+  return(vec)}
+
+## Pull P-Val and Pull F stat return the p-values and f-statstics from the list of output MultiPermanova puts out. 
+
+pullPval <- function(x){temp <- x$aov.tab$`Pr(>F`[1]; return(temp)}
+
+pullFstat <- function(x){temp <- x$aov.tab$`F.Model`[1]; return(temp)}
+
+
+
+## MultiPermanova: Function that takes an ordination method: Either t-SNE or MDS and produces a 
+
+MultiPermanova <- function(data, nperms = 5000, perp = 10, pca_options = FALSE){
+  
+  # initialize the size of the data and output lists
+  n <- nrow(data)
+  maplist <- list()
+  cmdlist <- list()
+  
+  #generate maps on holdout data
+  for(j in 1:n){
+    
+    #hold out a single row of the data (in the jth row)
+    #note - have to remove, as t-SNE doesn't like NAs
+    data_holdout <- data[-j,]
+    
+    #calculate distances FIRST
+    data_dist <- dist(data_holdout)
+    
+    #generate the classical MDS
+    mds1 <- cmdscale(data_dist, k = 2)
+    cmdlist[[j]] <- mds1
+    
+    #generate the t-SNE map and store x,y in map list
+    x <- Rtsne(data_dist, perplexity = perp, pca = pca_options, is_distance = TRUE)
+    maplist[[j]] <- x$Y
+  }
+  
+  
+  # Insert NAs into the holdout indices 
+  maplist_new <- list()
+  cmdlist_new <- list()
+  
+  
+  #
+  for(j in 1:length(maplist)){
+    #create temporary vectors with NAs for each of the XY coordingates
+    tempX <- insert(maplist[[j]][,1], j, NA)
+    tempY <- insert(maplist[[j]][,2], j, NA)
+    
+    tempcmdX <- insert(cmdlist[[j]][,1], j, NA)
+    tempcmdY <- insert(cmdlist[[j]][,2], j, NA)
+    
[…]

### 2021-08-21 PlotDevelopment.R

@@ -1,0 +1,125 @@
+## some live code development
+
+
+#x3 is a list of outputs from multipermanova
+out <- x3
+out$tsnelist
+
+
+library(tidyverse)
+library(ggplot2)
+library(ggridges)
+
+
+
+histlist1 <- list()
+histlist2 <- list()
+
+for(j in 1:length(out$tsnelist)){
+  histlist1[[j]] <- out$tsnelist[[j]]$f.perms %>% as_tibble() #%>% ggplot(aes(V1)) + geom_histogram(bins = 10)
+  histlist2[[j]] <-  out$cmdlist[[j]]$f.perms %>% as_tibble() #%>% ggplot(aes(V1)) + geom_density() + scale_x_log10()
+}
+
+
+## Left Panel (stacks the distributions on top of each other)
+
+df = histlist1 %>% bind_rows() %>% as_tibble()
+names(df) = "Density"
+df$Holdout = rep(1:length(histlist1),each = nrow(histlist1[[1]])) %>% as.factor()
+
+permplot = ggplot(df, aes(Density, color = Holdout)) + geom_density(alpha = 0.7) + 
+  scale_color_viridis_d(option = "B", end = 0.9) + guides(color = FALSE) + 
+   coord_flip() + theme_bw() + xlab("") # + ggtitle("Permutation Distribution") 
+
+permplot
+
+## Right Panel - Original Plot (shows the f-statistics)
+df2 = tibble(Holdout = 1:length(out$tsnelist), PValues = sapply(out$tsnelist, pullPval), Fstat = sapply(out$tsnelist, pullFstat))
+df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
+df2list[[j]] = df2
+
+#### TSNE Plots########################################################
+#Visualize the results and store in a list
+sigplot <- ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() #+ ggtitle("t-SNE: P-Values From Adonis")
+sigplot
+
+#Visualize the f stats and store in a list
+sigplotf <- ggplot(df2, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ylab("") + scale_color_viridis_d("Significant",option = "B", end = 0.8)#+ ggtitle("t-SNE: F Stats From Adonis")
+sigplotf
+
+library(ggpubr)
+
[…]

### 2021-08-21 Simulations_Influential_Points.R

@@ -1,0 +1,113 @@
+## Simulations
+
+#### Preliminary stuff
+source("OutlierCompFunctions.R")
+library(ggplot2)
+library(readr)
+library(tibble)
+library(dplyr)
+library(tidyr)
+library(janitor)
+library(magrittr)
+library(knitr)
+library(Rtsne)
+library(robustHD)
+library(GGally)
+library(stringr)
+library(plotly)
+library(R.utils)
+library(vegan)
+library(mnormt)
+
+plot_data_list <- list()
+plot_tsne_results_list <- list()
+plot_mds_results_list <- list()
+results_list <- list()
+
+for(j in 1:3){
+  
+  
+  #### Simulate the Data
+  set.seed(34536)
+  g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
+    t() %>%
+    data.frame()
+  g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(25, 20)) %>%
+    t() %>%
+    data.frame()
+  g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(25, 30)) %>%
+    t() %>%
+    data.frame()
+  
+  
+  yes_structure <- rbind(g1, g2, g3) %>%
+    data.frame() %>%
+    mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))
+  
+  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
+  
+  ## adds outliers in the last 3 indices (simulating from normal based on 5 sd's from mean)
+  add_outliers <- rbind(rnorm(10, 5, .5), rnorm(10, 5, .5), rnorm(10, 5, .5)) %>% as_tibble()
+  names(add_outliers) <- names(yes_structure_scale)
+  
+  
+  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
+  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))
+  
+  
+  #Visualize the data and store in a list
+  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In")
+  plot_data_list[[j]] <- datplot
+  
+  
+  
+  #### Run the Permanova mapping
+  
+  
[…]

### 2021-08-21 OldOutlierCompFunctions.R
adding additional writeup files to compfunctoin files

@@ -1,6 +1,3 @@
-## Script to functionalize this process
-
-
 # preliminary functions: 
 
 ## Vectorize Dist: Function that takes a dataframe and returns a vectorized version of a distance matrix applied to that data matrix. 
@@ -18,15 +15,9 @@ pullPval <- function(x){temp <- x$aov.tab$`Pr(>F`[1]; return(temp)}
 pullFstat <- function(x){temp <- x$aov.tab$`F.Model`[1]; return(temp)}
 
 
+### Old multipermanova function
 
-### Dependent Functions
-
-#generate maps on holdout data (jth iteration is held out)
-# this function should be able to operate on multiple mapping techniques or a single one 
-
-## MultiPermanova: Function that takes an ordination method: Either t-SNE or MDS and produces a 
-
-Generate_Holdout_Maps <- function(data, j, maptype = "MDS",...){  
+Generate_Holdout_Maps <- function(data, j, ...){  
   
   #hold out a single row of the data (in the jth row)
   #note - have to remove, as t-SNE doesn't like NAs
@@ -35,61 +26,39 @@ Generate_Holdout_Maps <- function(data, j, maptype = "MDS",...){
   #calculate distances FIRST
   data_dist <- dist(data_holdout)
   
-  #### Creates Maps (based on several methods: tSNE and MDS currently)
+  #### Creates Maps (based on several methods)
   
-  if(maptype %in% c("MDS", "mds")){
-    #generate the classical MDS
-    map <- cmdscale(data_dist, k = 2)
-  }
-  else if(maptype == "tSNE"){
-    #generate the t-SNE map and store x,y in map list
-    map <- Rtsne(data_dist, perplexity = perplexity, pca = pca_options, is_distance = TRUE)$Y
-  }
-  else{print("No maptype selected.")}
+  #generate the classical MDS
+  mds1 <- cmdscale(data_dist, k = 2)
   
+  #generate the t-SNE map and store x,y in map list
+  x <- Rtsne(data_dist, perplexity = perplexity, pca = pca_options, is_distance = TRUE)
   
-  return(list(map = map))
+  return(list(mds = mds1, tsne = x$Y))
 }
 
 
 
-## Generates the Permanova test 
-# 1: listobject: list of maps (maplist, cmdlist) to evaluate
-# 2: nperms - should inherit from above
-# 3: j: index to be looped through 
[…]

### 2021-08-21 OutlierCompNew.R

@@ -1,30 +1,8 @@
-## Script to functionalize this process
+## Draft version of outlier comparison function agnostic to mapping technique
 
-
-# preliminary functions: 
-
-## Vectorize Dist: Function that takes a dataframe and returns a vectorized version of a distance matrix applied to that data matrix. 
-## This function is used for generating vectorized data when we step through the Mantel test. 
-
-vectorize_dist <- function(df) {
-  tmp <- dist(df)
-  vec <- c(tmp)
-  return(vec)}
-
-## Pull P-Val and Pull F stat return the p-values and f-statstics from the list of output MultiPermanova puts out. 
-
-pullPval <- function(x){temp <- x$aov.tab$`Pr(>F`[1]; return(temp)}
-
-pullFstat <- function(x){temp <- x$aov.tab$`F.Model`[1]; return(temp)}
-
-
-
-### Dependent Functions
 
 #generate maps on holdout data (jth iteration is held out)
 # this function should be able to operate on multiple mapping techniques or a single one 
-
-## MultiPermanova: Function that takes an ordination method: Either t-SNE or MDS and produces a 
 
 Generate_Holdout_Maps <- function(data, j, maptype = "MDS",...){  
   
@@ -53,24 +31,8 @@ Generate_Holdout_Maps <- function(data, j, maptype = "MDS",...){
 
 
 
-## Generates the Permanova test 
-# 1: listobject: list of maps (maplist, cmdlist) to evaluate
-# 2: nperms - should inherit from above
-# 3: j: index to be looped through 
-# HDpermanova <- function(listobject, nperms, j){
-#   #generate the indicator variable for each feature
-#   indicator_df <- rep(0, nrow(listobject[[1]])) %>% as.data.frame()
-#   
-#   names(indicator_df) <- "Indicator"
-#   indicator_df$Indicator[j] = 1
-#   
-#   #each item in the list is an Adonis test using the indicator for the holdout variable
-#   #note that this is set to do parallel processing with 2 cores
-#   model_output <- adonis(cordist ~ Indicator, data = indicator_df, permutations = nperms, parallel = 2)
-#   
-#   return(list(out = model_output)) 
-# }
 
+#mapping method supported
 
 MultiPermanova <- function(data, nperms = 5000, perplexity = 10, pca_options = FALSE, maptype = "MDS"){
   
[…]

### 2021-08-21 PaperDraft.Rmd
adds initial draft of paper

--- 2021-08-21 FormalMethodologyWriteup.Rmd
@@
 ---
-title: "Formal Writeup"
+title: "Formal Paper Draft"
+subtitle: "Influence Diagnostics for High-Dimensional Ordination Techniques"
 author: "Paul Harmon"
-date: "7/25/2021"
+date: "8/21/2021"
 output: html_document
@@
 library(mnormt)
-
-Generate_Holdout_Maps <- function(data, j, ...){
-
-  #hold out a single row of the data (in the jth row)
-  #note - have to remove, as t-SNE doesn't like NAs
-  data_holdout <- data[-j,]
-
-  #calculate distances FIRST
-  data_dist <- dist(data_holdout)
-
-  #### Creates Maps (based on several methods)
-
-  #generate the classical MDS
-  mds1 <- cmdscale(data_dist, k = 2)
-
-  #generate the t-SNE map and store x,y in map list
-  x <- Rtsne(data_dist, perplexity = perplexity, pca = pca_options, is_distance = TRUE)
-
-  return(list(mds = mds1, tsne = x$Y))
-}
-
-perplexity = 10
-
 ```
@@
 
-## Goal
+# High-dimensional Data Visualization
+## What is t-SNE
+## Problems with using t-SNE
+## Whitening/Pre-PCA
 
-The goal of this method is to develop a diagnostic that can be used to help identify influential points and potentially quantify how much they impact the resulting mapping, irrespective of how that mapping was created.
-Additionally, this method can be used to assess the sensitivity of each method to influential points and the impact that sensititivy might have on how an end user might want to interpret the resulting map produced.
-Note that while robustness to influential points might at first seem like a positive trait for an ordination method, masking distinctly different observations in a mapping may have negative consequences as well.
+# Other Multivariate Dimension-Reduction Tools
+## MDS
+## Sammon Mapping
+## UMAP
 
-This document overviews the statistical methodology we have developed to identify influential points that, when excluded from a dataset, can meaningfully impact the shape of an ordination created from data.
-We overview our methodology step by step.
+# Defining “Influence” in High Dimensional Maps
+## Influence in Regression - Cook’s Distance
+## Influence in
 
 # Methods
-
-This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE.
-In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.
-
-Our Permanova-based approach is as follows:
-
-1.
-For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.
-(For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA).
-2. Generate a distance matrix of the nx1 by 2 matrix. Do this for each of the held-out observations. 
-3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
-4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-|\sqrt{}\. 
-5.
-Apply non-parametric PERMANOVA tests for each holdout observation - this means that on a 100-observation dataset, you apply 100 individual tests where each test uses an independent variable that is equal to 1 for the index of the holdout observation and 0 elsewhere.
-The hypotheses are specified as such:
-
-+ $H_0$: No differences in map with observation j removed vs. those without j removed 
-+ $H_a$: Some difference in map with observation j removed
-
-We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Benjamini-Hochberg False Discovery Rate or Bonferroni corrections (or something similar).
-We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth adonis test would be small.
-
-# Assessment on Data
-
-## Token Example
-
-We first step through an example of the method as utilized on a relatively small dataset, to show how the algorithm works.
-Let's consider an example with 63 observations simulated from a multivariate normal distribution.
-
-The code below simulates the data with some clearly abberant values.
-
-```{r simulate_token_data, include = TRUE, message = FALSE, warning = FALSE}
-set.seed(34536)
-g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
-  t() %>%
-  data.frame()
-g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(25, 20)) %>%
-  t() %>%
-  data.frame()
-g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(25, 30)) %>%
-  t() %>%
-  data.frame()
-
-yes_structure <- rbind(g1, g2, g3) %>%
-  data.frame() %>%
-  mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))
-
-yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
-
-## adds outliers in the last 1 index (simulating from normal based on 5 sd's from mean)
-add_outliers <- rbind(rnorm(10, 5, .5)) %>% as_tibble()
-names(add_outliers) <- names(yes_structure_scale)
-
-yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
-yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
-```
-
-1.
-For each observation in the dataset, remove it, and generate a (t-SNE or MDS) map in 2 dimensions from the resulting (n-1) observations.
-(For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA).
-The code below shows this as it pertains to the "multipermanova" function that I have written - it makes use of a function called "insert" from R.utils that adds NAs for each index that is held out.
-
-```{r, eval = TRUE, message = FALSE}
-#initializes some stuff 
-data = yes_structure_final
-n <- nrow(data)
-  maplist <- list()
-  cmdlist <- list()
-
-#generate maps on holdout data
-  for(j in 1:n){
-
-    temp <- Generate_Holdout_Maps(data, j)
-    cmdlist[[j]] <- temp$mds
-    maplist[[j]] <- temp$tsne
-
-  }
-
-# Insert NAs into the holdout indices 
-  maplist_new <- list()
-  cmdlist_new <- list()
-
-#replaces holdouts with NAs
-for(j in 1:length(maplist)){
-    #create temporary vectors with NAs for each of the XY coordingates
-    tempX <- R.utils::insert(maplist[[j]][,1], j, NA)
-    tempY <- R.utils::insert(maplist[[j]][,2], j, NA)
-
-    tempcmdX <- R.utils::insert(cmdlist[[j]][,1], j, NA)
-    tempcmdY <- R.utils::insert(cmdlist[[j]][,2], j, NA)
-
-    maplist_new[[j]] <- tibble(X = tempX, Y = tempY)
-    cmdlist_new[[j]] <- tibble(X = tempcmdX, Y = tempcmdY)
-}
-
-#prints out the maps
-maplist_new[[1]] %>% head()
-
-plot(maplist_new[[1]], pch = 20, main = "Example of maplist (t-SNE), Observation 1 Held Out")
-
-```
-
-2. Generate a distance matrix of the n by 2 matrix. Do this for each of the held-out observations.
-
-3. Vectorize the distance matrix and generate a correlation matrix from the combination of these vectorized distances.
-
-We vectorize distance matrices in the following way, where $d12$ refers to the distance between the first and second observations (in the map).
-I'm ignoring the upper/lower triangle symmetry in this example.
-As an example, here's what this looks like in a very simplistic 3 observation case.
-
-$$\mathbf{Dist(X_{nxn})} = \left[\begin{array}
-{rrr}
-NA & NA & NA \\
-NA & 0 & d_{23} \\
-NA & d_{32} & 0 
-\end{array}\right]
-$$
-
-The vectorized form (shown as a row vector) would be: 
-$$\mathbf{c(Dist(X))^T} = \left[\begin{array}
-{r}
-NA & NA & NA &
-NA & 0 & d_{32} &
-NA & d_{23} & 0
-\end{array}\right]
-$$
-Then, the correlation matrix as follows, where each column comes from the ith holdout map.
-Note that the below notation reads: $d_{23, 1}$ is the distance between observations 2 and 3 on the map that holds out the 1st observation.
-
-$$ \mathbf{Cor(Dist(X))} = \left[\begin{array}
-{rrr}
-NA & d_{12, 2} & d_{13,3} \\
-NA & NA & d_{23,3} \\
-NA & d_{31,2} & NA \\
-NA & NA & ...\\
-0 & NA & .\\
-d_{32,1} & NA & .\\
-NA & d_{13,2} & .\\
-d_{23,1} & NA & .\\
-0 & 0 & .\\
-\end{array}\right]
-$$
-
-Because of the holdout NAs, we do not have complete cases to compare so instead we use pairwise-complete observations.
-We can see a visual representation of the NAs for a sample of the distance matrices below:
-
-```{r}
-#vectorizes the distances and outputs a list of vectorized distance matrices 
-  distmat_list <- lapply(maplist_new, dist)
-  distmat_list_cmd <- lapply(cmdlist_new, dist)
-
-levelplot(as.matrix(distmat_list[[1]]), main = paste0("NA Pattern for ",1, "st Holdout"))
-levelplot(as.matrix(distmat_list[[3]]), main = paste0("NA Pattern for ",3, "rd Holdout"))
-levelplot(as.matrix(distmat_list[[5]]), main = paste0("NA Pattern for ",5, "th Holdout"))
-
-```
-
-The code chunk below shows both steps simultaneously.
-Note that generating the pearson correlation matrix from the vectorized distance matrices is done using pairwise complete observations - the NAs we include from the holdout set can cause potential problems when this is not specified.
-
-```{r, eval = TRUE}
- #vectorizes the distances and outputs a list of vectorized distance matrices 
-  dist_list <- lapply(maplist_new, vectorize_dist)
-  dist_list_cmd <- lapply(cmdlist_new, vectorize_dist)
-
-  ## takes the vectorized distance matrices and smashes them into a single data frame - to build correlation matrix
-
-  dist_mat <- dist_list %>% as_tibble(.name_repair = "minimal")
-  dist_mat_cmd <- dist_list_cmd %>% as_tibble(.name_repair = "minimal")
-
-  #prints out the data frame of vectorized distance matrices to show what we're looking at
-  head(dist_mat)
-
-  #dimensions are n*(n-1)/2 since the distance matrix is triangular - see below
-  dim(dist_mat)
-
-```
-
-We calculate the correlation matrix from the above:
-
-```{r}
-
-  #Generates the correlation matrix based on pairwise complete observations, and pearson correlation
-
-  #tsne
-  cormat <- cor(dist_mat, use = "pairwise.complete.obs", method = "pearson")
-  cormat %>% dim()
-
-  #cmd
-  cormat_cmd <- cor(dist_mat_cmd, use = "pairwise.complete.obs", method = "pearson")
-  cormat_cmd %>% dim()
-
-```
-
-4.
-Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using the result that $D_{ij} = \sqrt{2(1-cor)}$.
-Note that here, we expect that all the correlation values are greater than 0 - and as shown, they are.
-
-*Note also that with this new distance matrix, we can actually create 2D visualizations (maps) from the distance matrix, creating a "map-of-maps" to show which of the ordinations are different from the rest.
-(Similarly, you could cluster maps or perform any distance-based method on this matrix to explore differences in the holdout ordinations).*
-
-```{r}
-## sanity check - these are all greater than 0
-cormat %>% min()
-##generate distance on correlation matrix
-  cordist <- sqrt(2*(1-cormat)) %>% as.dist()
-  cordist_cmd <- sqrt(2*(1-cormat_cmd)) %>% as.dist()
-```
-
-5.
-Apply non-parametric PERMANOVA tests for each holdout observation, using an indicator that is 1 for the holdout observation and 0 for all others.
-Cycle the indicator for each index - so you will apply $n$ individual tests.
-
-In the code below, the data does not change - the calls to 'adonis' change only in that the index of the "1" for the indicator variable chanegs at each iteration.
-Thus, the indicator for the first holdout observation is [1,0,0,...] and so on and so forth.
-
-Note that for adonis, this version uses 1000 permutations rather than a permutation matrix, meaning that this version re-samples the $n+1$ different unique permutations of the indicator which has a single $1$ and $0$'s in all the other indices.
-*We are working on being a bit more efficient with this permutation strategy so that we do not sample more than necessary*.
-
-```{r}
-  nperms = 1000
-  model_list <- list()
-  model_list_cmd <- list()
-
- for(j in 1: nrow(as.matrix(cordist))){
-    #generate the indicator variable for each feature
-    indicator_df <- rep(0, nrow(maplist_new[[1]])) %>% as.data.frame()
-    indicator_df_cmd <- rep(0, nrow(cmdlist_new[[1]])) %>% as.data.frame()
-
-    names(indicator_df) <- names(indicator_df_cmd) <- "Indicator"
-    indicator_df$Indicator[j] = 1
-    indicator_df_cmd$Indicator[j] = 1
-
-     ## create the permutation matrix
-  #initialize
-permat1 <- matrix(0, nrow = length(indicator_df$Indicator), ncol = length(indicator_df$Indicator))
-permat1[1,] <- 1:length(indicator_df$Indicator)
-
-for(i in 2:nrow(permat1)){
-  permat1[i,] <- c(i:length(indicator_df$Indicator), 1:(i-1))
-}
-
-    #each item in the list is an Adonis test using the indicator for the holdout variable
-    model_list[[j]] <- adonis(cordist ~ Indicator, data = indicator_df, permutations = permat1, parallel = 2)
-    model_list_cmd[[j]] <- adonis(cordist_cmd ~ Indicator, data = indicator_df_cmd, permutations =  permat1)
-
-  }
-
-```
-
-Note that in the Adonis function, we do obtain a few negative f-statistic values in certain cases.
-These are the null values based on the permutations used.
-This seems to occur only for the MDS data - in the cases where we do see negative values, they are quite close to 0.
-
-```{r}
-
-#from t-SNE
-model_list[[1]]
-
-model_list[[1]]$f.perms %>% min()
-model_list[[1]]$f.perms %>% hist(main = "Example of Permutation Distribution: t-SNE")
-
-#from cmd - an example with negatives included (note that they are all the same value )
-model_list_cmd[[1]]$f.perms %>% min()
-model_list_cmd[[1]]$f.perms[which(model_list_cmd[[1]]$f.perms<0)]
-model_list_cmd[[1]]$f.perms %>% hist(main = "Example of Permutation Distribution: MDS")
-
-```
-
-# Appendix
-
-## Outliers and Influential Points
-
-Similarly, the resulting maps produced by ordination methods like MDS, t-SNE and the like can be impacted by the inclusion or exclusion of points.
-In short, an "influential" point in a lower dimensional mapping can be defined as one that, when excluded, drives meaningful differences in the ordination method that lead to a different shaped map as compared to when that point is included.
-
-Difference between outliers and influential points.
-
-**talk about influence functions** (Joliffe and cited papers)
-
-Joliffe (2002) overviews tools for identifying influential observations in Pricipal Component Analysis, which is a related dimension-reducting technique to MDS (at least in a Euclidean setting).
-However, the literature on more recently-developed methods lacks a similar metric, most notably one that can be used or compared across different mapping methods.
-
-This section overviews some work into identifying multivariate outliers that we've looked into.
-We have not found much in the way of users applying Mantel tests or Permanova-type tests on t-SNE or Classical MDS results in the way that we are thinking about things.
-In fact, much of the literature around robustness and outliers centers on dealing with the problem by modifying the dimension reduction technique rather than identifying it.
-
-Blouvshtein and Cohen-Or (2018) analyzed the effect of outliers on the maps produced by classical multidimensional scaling methods.
-In general, these methods are not particularly robust to the effects of multivariate outlier.
-They propose a method to detect outliers prior to the generation of the MDS map so that they can be removed from the data.
-Their method involves treating the distances input to the algorithm as edges of a complete graph that connects each data point.
-
-These edges $d1, d2, d3$ can then be used to form triangles - in general, outliers tend to *break* the triangle inequality and inliers tend to satisfy it.
-They recommend a rule to examine the histogram H of $b$, the number of edges of broken triangles, and generate a threshold $\phi$ that is intended to identify inliers vs.
-outliers.
-
-Two other papers, including Forrero and Giannakis (2012) as well as Kong et al.
-(2019) frame the problem in a slightly different way.
-They consider a penalized *stress* metric for optimization, utilizing regularization to better identify outlier points.
-In both cases, they are able to identify outliers using a majorization-minimization procedure that iteratively produce more outlier-robust maps of the lower-dimensional embedding of interest.
+## Sparse t-SNE
+## Cross Validation and Mantel-esque Test

### 2021-08-21 PaperDraft.Rmd
adding changes to some docs

@@
 
-Despite those differences, tools like t-SNE and MDS are succeptible to abberant data values, much in the same way that a regression model might be succeptible to outliers and influential points.
+Despite those differences, tools like t-SNE and MDS are susceptible to aberrant data values, much in the same way that a regression model might be susceptible to outliers and influential points.
 In a regression model, the estimated coefficients or predictions might be affected in meaningful fashion by the inclusion or exclusion of a single point - such points are referred to as influential (Cook, 1977).
@@
 
-# High-dimensional Data Visualization
-## What is t-SNE
+The goal of this method is to develop a diagnostic that can be used to help identify influential points and potentially quantify how much they impact the resulting mapping, irrespective of how that mapping was created.
+Additionally, this method can be used to assess the sensitivity of each method to influential points and the impact that sensititivy might have on how an end user might want to interpret the resulting map produced.
+Note that while robustness to influential points might at first seem like a positive trait for an ordination method, masking distinctly different observations in a mapping may have negative consequences as well.
+
+This document overviews the statistical methodology we have developed to identify influential points that, when excluded from a dataset, can meaningfully impact the shape of an ordination created from data.
+We overview our methodology step by step.
+
+# High-dimensional Data Visualization with t-SNE
+
+One of the notable new methods for high-dimensional data visualization is t-distributed Stochastic Neighbor Embedding, or t-SNE (Van der Maaten, 2011).
+Since the time that Laurens Van der Maaten published the first t-SNE paper in 2011, it has been cited more than 21,000 times in use cases ranging from biology (Amir et al, 2013) and genetics () to economics (https://economics.yale.edu/sites/default/files/files/Faculty/Tsyvinski/tSNE_draft15.pdf) to natural language processing (https://aneesha.medium.com/using-tsne-to-plot-a-subset-of-similar-words-from-word2vec-bb8eeaea6229) and machine learning.
+The method's prevalence is astounding - it is used not only in academia but in industry problems, including tools like FoodGenius (https://github.com/foodgenius/tsneplot).
+
+The method centers on calculation of a balance between two similarity metrics.
+The first, called $p_{ij}$,is a conditional probability based on the Euclidean distance between pairs of points in the high-dimensional space.
+In the lower-dimensional representation, this similarity is defined as $q_{ij}$.
+
+A good representation of the high-dimensional space should mean that $p_{ij}$ and $q_{ij}$ look reasonably similar to each other; since these are both simply joint probabilities, the objective function of t-SNE is simply a symmetricized Kullbeck-Leibler divergence between the two, defined below:
+
+$$C = \sum_{i} \sum_{j} p_{j|i} log(\frac{p_{j|i}}{q_{j|i}})$$
+
+t-SNE is an improvement over Stochastic Neightbor Embedding (SNE), which makes use of Gaussian distributions to model both $p$ and $q$ make preserving local structure somewhat difficult.
+Most of the minimization of objective functions is done via gradient descent (and some methods require simulated annealing because the algorithm can get caught in local optima).
+Critically, because the optimization is non-convex, and as a consequence of the gradient descent methods used to optimize it, t-SNE has the potential to provide different maps when run on the same data at the same perplexity.
+
 ## Problems with using t-SNE
-## Whitening/Pre-PCA
+
+Although t-SNE is highly effective at creating 2 (or 3) dimensional ordinations of high-dimensional data, the mathematical
+
+Finally, performing t-SNE on large datasets can be computationally slow, necessitating a random-walk version that is discussed in a later paper by van der maaten (2014).
+
+## Reduction of Initial Dimensionality with PCA
+
+Additionally, t-SNE is subject to a pre-processing step involving principal component analysis (PCA).
+In most implementations, "whitening" via PCA is performed on the initial dataset (or input distance matrix) to reduce the dimensionality to a reasonable value.
+For instance, in Van der Maaten's initial paper, several examples reduced the initial dimensionality of the datato 30 for computational gains (Van der maaten, 2011).
+
+Depending on the size of the initial dimensionality of the data, this step may drastically reduce the number of dimensions that t-SNE needs to deal with.
+However, it has some potential to impact the resulting ordination, particularly if the number of initial dimensions kept is relatively low or PCA does not do a great job of fitting the original dataset.
+At best, the t-SNE mapping is only as good as the initial PCA is.
+For the simulations shown in this paper, no whitening was performed, and the original distance matrix was passed to t-SNE directly.
 
@@
 # Defining “Influence” in High Dimensional Maps
+
 ## Influence in Regression - Cook’s Distance
+
+In regression models, the concept of influential points is
+
+## Sparse t-SNE
+## Cross Validation and Mantel-esque Test
+
 ## Influence in
@@
 # Methods
-## Sparse t-SNE
-## Cross Validation and Mantel-esque Test
+
+This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE.
+In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.
+
+Our Permanova-based approach is as follows:
+
+1.
+For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.
+(For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA).
+2. Generate a distance matrix of the nx1 by 2 matrix. Do this for each of the held-out observations. 
+3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
+4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-|\sqrt{}\. 
+5.
+Apply non-parametric PERMANOVA tests for each holdout observation - this means that on a 100-observation dataset, you apply 100 individual tests where each test uses an independent variable that is equal to 1 for the index of the holdout observation and 0 elsewhere.
+The hypotheses are specified as such:
+
++ $H_0$: No differences in map with observation j removed vs. those without j removed 
++ $H_a$: Some difference in map with observation j removed
+
+We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Benjamini-Hochberg False Discovery Rate or Bonferroni corrections (or something similar).
+We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth adonis test would be small.
+
+# Results
+
+# Discussion
+
+t-SNE is not particularly sensitive to single points that are highly dissimilar to others in the dataset.
+It is often not obvious that such observations are present when viewing a t-SNE map.
+In the case of other methods, like MDS, influential points change the configuration of the ordination significantly.

### 2021-08-21 tSNE Introduction Document.Rmd
adding changes to some docs

@@
 ## How is t-SNE used
+
+One of the notable new methods for high-dimensional data visualization is t-distributed Stochastic Neighbor Embedding, or t-SNE (Van der Maaten, 2011).
+Since the time that Laurens Van der Maaten published the first t-SNE paper in 2011, it has been cited more than 21,000 times in use cases ranging from biology (Amir et al, 2013) and genetics () to economics (https://economics.yale.edu/sites/default/files/files/Faculty/Tsyvinski/tSNE_draft15.pdf) to natural language processing (https://aneesha.medium.com/using-tsne-to-plot-a-subset-of-similar-words-from-word2vec-bb8eeaea6229) and machine learning.
+The method's prevalence is astounding - it is used not only in academia but in industry problems, including tools like FoodGenius (https://github.com/foodgenius/tsneplot).
+
+The method centers on calculation of a balance between two similarity metrics.
+The first, called $p_{ij}$,is a conditional probability based on the Euclidean distance between pairs of points in the high-dimensional space.
+In the lower-dimensional representation, this similarity is defined as $q_{ij}$.
+
+A good representation of the high-dimensional space should mean that $p_{ij}$ and $q_{ij}$ look reasonably similar to each other; since these are both simply joint probabilities, the objective function of t-SNE is simply a symmetricized Kullbeck-Leibler divergence between the two, defined below:
+
+$$C = \sum_{i} \sum_{j} p_{j|i} log(\frac{p_{j|i}}{q_{j|i}})$$
+
+t-SNE is an improvement over Stochastic Neightbor Embedding (SNE), which makes use of Gaussian distributions to model both $p$ and $q$ make preserving local structure somewhat difficult.
+Most of the minimization of objective functions is done via gradient descent (and some methods require simulated annealing because the algorithm can get caught in local optima).
+Critically, because the optimization is non-convex, and as a consequence of the gradient descent methods used to optimize it, t-SNE has the potential to provide different maps when run on the same data at the same perplexity.
 

### 2021-08-22 PaperDraft.Rmd
mas changeados

@@
 title: "Formal Paper Draft"
-subtitle: "Influence Diagnostics for High-Dimensional Ordination Techniques"
 author: "Paul Harmon"
 date: "8/21/2021"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
+subtitle: Influence Diagnostics for High-Dimensional Ordination Techniques
 ---
@@
 
-# High-dimensional Data Visualization with t-SNE
+# High-dimensional Data Visualization Tools
+
+## t-SNE
 
@@
 
-## Problems with using t-SNE
-
-Although t-SNE is highly effective at creating 2 (or 3) dimensional ordinations of high-dimensional data, the mathematical
-
-Finally, performing t-SNE on large datasets can be computationally slow, necessitating a random-walk version that is discussed in a later paper by van der maaten (2014).
-
-## Reduction of Initial Dimensionality with PCA
+### Reduction of Initial Dimensionality with PCA
 
@@
 
-# Other Multivariate Dimension-Reduction Tools
-## MDS
-## Sammon Mapping
-## UMAP
+## Multidimensional Scaling
+
+Classical multidimensional scaling (MDS) is closely related to PCA (an in some cases, equivalent) (Gower, 1966).
+MDS takes as input a distance or dissimilarity matrix representing the distance between pairs of observations - this distance need not simply be Euclidean.
+Classical multidimensional scaling returns maps in teh same metric as the original distance or dissimilarity matrix.
+
+## Sparse t-SNE
+## Cross Validation and Mantel-esque Test
 
@@
 
-In regression models, the concept of influential points is
+In regression models, an outlier is typically denoted as a point that is either much larger or smaller than similar observations on a specific scale.
+However, an outlier need not be an influential point.
+An influential point simply needs to be outlying enough relative to similar observations that it impacts the fit of the regression line.
+the identification of influential points that can impact the results of a model is commonplace.
 
-## Sparse t-SNE
-## Cross Validation and Mantel-esque Test
+Cook (1977) proposed a measure of influence called "Cook's Distance" that has since become one of the most common tools for diagnosing influential points and outliers.
+The idea is predicated on the concept that by leaving an out an influential observation, the results of the model should change.
+This process of jacknifing observations is done for all the observations in the data.
 
-## Influence in
+Cook's distance is defined as follows for the $ith$ observation, with$\hat{j}_{j}$ the fitted response value for the model withthe $ith$ observation omitted, $s^2$ defined as the mean-squared error for the model and $p$ the number of variables for each observation:
+$$ D_i = \frac{\sum_{j=1}^{n} (\hat{y_j} - \hat{y}_{j(i)})^2}{ps^2}$$
 
-# Methods
+Although Cook's distance is essentially an F statistic from an $F_{p, n-p}$ distribution, and the method does not use p-values to determine how "influential" a point is.
+Rather, cutoff values are typically assessed, or observations with relatively large Cook's distance values are considered for investigation of influence.
 
-This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE.
-In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.
+```{r}
+#image of an influential point
+x = rnorm(100, 10, 3)
+y = 3 + 2*x + rnorm(100, 0, 1)
 
-Our Permanova-based approach is as follows:
+lm1 = lm(y~x)
+par(mfrow = c(2,2))
+plot(lm1)
+
+```
+
+## Outliers and Influential Points in Maps
+
+Similar to a regression setting, the resulting maps produced by ordination methods like MDS, t-SNE and the like can be impacted by the inclusion or exclusion of points.
+In short, an "influential" point in a lower dimensional mapping can be defined as one that, when excluded, drives meaningful differences in the ordination method that lead to a different shaped map as compared to when that point is included.
+An outlier, on the other hand, might look different in the high-dimensional space (or on a subset of features), but would not meaningfully change the shape of the map when left out.
+
+Joliffe (2002) overviews tools for identifying influential observations in Pricipal Component Analysis, which is a related dimension-reducing technique to MDS.
+Influence functions (Critchley, 1985; Hampel, 1974; Huber, 1981) can be defined for regression methods as well as for loadings in PCA.
+However, the literature on more recently-developed methods lacks a similar metric, most notably one that can be used or compared across different mapping methods.
+
+Much of the literature around robustness of mapping results and multivariate outliers centers on dealing with the problem by modifying the dimension reduction technique rather than identifying it.
+Blouvshtein and Cohen-Or (2018) analyzed the effect of outliers on the maps produced by classical multidimensional scaling methods.
+In general, these methods are not particularly robust to the effects of multivariate outlier.
+They propose a method to detect outliers prior to the generation of the MDS map so that they can be removed from the data.
+Their method involves treating the distances input to the algorithm as edges of a complete graph that connects each data point.
+
+These edges $d1, d2, d3$ can then be used to form triangles - in general, outliers tend to *break* the triangle inequality and inliers tend to satisfy it.
+They recommend a rule to examine the histogram H of $b$, the number of edges of broken triangles, and generate a threshold $\phi$ that is intended to identify inliers vs.
+outliers.
+
+Two other papers, including Forrero and Giannakis (2012) as well as Kong et al.
+(2019) frame the problem in a slightly different way.
+They consider a penalized *stress* metric for optimization, utilizing regularization to better identify outlier points.
+In both cases, they are able to identify outliers using a majorization-minimization procedure that iteratively produce more outlier-robust maps of the lower-dimensional embedding of interest.
+
+# Method Overview
+
+We instead propose a method for identifying influential observations that is agnostic to the mapping method.
+In addition, our method follows a philosophical framework similar in flavor to Cook's distance for mapping methods; however, it relies on a different set of tools than its regression analogue.
+
+Similar to Cook's Distance, we propose calculating hold-out maps, with each of the 1...n observations held out singly.
+Rather than basing the test on a standard F distribution, as is Cook's distance, our method utilizes permutation-based MANOVA (Permanova) tests to calculate pseudo-F test statistics.
+Our method borrows from some of the machinery of the mantel test (Mantel, 1967), which is a method for comparing correlation between two matrices.
+
+The proposed approach is as follows:
 
@@
 3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
-4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-|\sqrt{}\. 
+4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-\|\sqrt{stuff}$. 
 5.
@@
 
-# Results
+# Simulation
+
+To test the method
+
+# Overview of Real Data Results
+
+## Penguin Data
+
+The Palmer Penguin dataset (Horst et al., 2020) contains information pertaining to three different species of penguins.
+The dataset is intended to be an alternative to the ever-popular iris dataset (cite) as each of the penguins form relatively distinct clusters when measured on a series of biometric markers.
+
+## Carnegie Classification
+
+The Carnegie Classification (CC) is a method for classifying institutions of higher education based on several metrics of institutional productivity.
+The 2015 version of CC was based on data ranging from doctorates awarded in STEM, Non-STEM, Humanities, Social Sciences, and Other (professional) fields, as well as research expenditures in STEM and Non-STEM fields and two proxies for institutional size (tenurable faculty headcount and research faculty headcount).
+
+The Carnegie Classification is typically based on a scaled version of the data due to the presence of outlier institutions on several metrics.
+In particular, some institutions spend a great deal of expenditures on research compared to others, so the actual CC is based on two indices generated using Principal Component Analysis on the ranked data (Cite Harmon, Kosar).
+
+For this analysis, we instead look at scaled versions of the unranked data in part to assess whether or not some of the relative extreme points appear as influential relative to other universities in the group.
 

### 2021-08-22 PaperDraft.Rmd
kmw

@@
 
+Overview of Penguin data.
+
+Additionally, we simulated an additional penguin that stands out significantly from the additional penguins.
+The penguin is 5 standard deviations above or below the average for each of the measurements.
+
 ## Carnegie Classification
@@
 
+```{r}
+CC2015 <- read.csv("https://raw.githubusercontent.com/paulharmongj/Carnegie_SEM/master/data/CC2015data.csv")
+
+CC2015_filter <- CC2015 %>% filter(BASIC2015 %in% c(15,16,17)) %>% select(PDNFRSTAFF, S.ER.D, NONS.ER.D, FACNUM, HUM_RSD, STEM_RSD, OTHER_RSD, SOCSC_RSD)
+perplexity = 30
+
+## NEED TO STANDARDIZE THE DATA FIRST (MAKE UNITLESS)
+CCscale <- CC2015_filter %>% lapply(scale) %>% as_tibble()
+
+out <- MultiPermanova(CCscale, maptype = "MDS")
+```
+
 # Discussion

### 2021-09-21 FormalMethodologyWriteup.Rmd
adding additional writeup files to compfunctoin files
adds initial draft of paper
adding updates to simulation docs

@@
 
-perplexity = 10
-
 ```
@@
 
-Consider the problem of visualizing multivariate (or potentially high-dimensional) data.
-In many cases, it is necessary to generate a 2 or 3-D mapping of some higher-dimensional space, projecting the high-dimensional data into an ordination that can be printed on paper.
+Consider the problem of visualizing multivariate data (or potentially multivariate in many or high-dimensions).
+In many cases, it is useful to generate a 2 or 3-D mapping of some higher-dimensional space, projecting the high-dimensional data into an ordination that can be directly visualized.
 There are many methods that exist to accomplish this task, ranging from tools such as classicial multidimensional scaling (MDS) to more modern tools like t-distributed Stochastic Neighbor Embedding.
@@
 Unlike regression or classification, most tools for dimension reduction and higher-dimensional data visualization are unsupervised, meaning that they do not take advantage of some marked response variable; rather, they capitalize on the signal present in a set of features and produce a data-driven result.
+In regression models, we have a suite of tools to measure and assess unusual observations related to response vs.
+model predictions.
 
-Despite those differences, tools like t-SNE and MDS are succeptible to abberant data values, much in the same way that a regression model might be succeptible to outliers and influential points.
 In a regression model, the estimated coefficients or predictions might be affected in meaningful fashion by the inclusion or exclusion of a single point - such points are referred to as influential (Cook, 1977).
 A suite of tools have been developed to identify and measure the impact of influential points on the results of regression models.
+Tools like t-SNE and MDS are also succeptible to abberant data values, much in the same way that a regression model might be succeptible to outliers and, specifically, influential points.
 
@@
 
-The goal of this method is to develop a diagnostic that can be used to help identify influential points and potentially quantify how much they impact the resulting mapping, irrespective of how that mapping was created.
+The goal of this method is to develop a diagnostic that can be used to help identify influential points and potentially quantify how much they impact the resulting mapping, and create a tool that can be used with any mapping technique.
 Additionally, this method can be used to assess the sensitivity of each method to influential points and the impact that sensititivy might have on how an end user might want to interpret the resulting map produced.
@@
 This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE.
-In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.
+In general, we seek to identify either **multivariate outliers** (or groups of them) which, for some reason, cause large differences in the shape of the resulting ordination.
+The methods rely on a few different ideas.
+First, PERMANOVA (Anderson, 2001) is a non-parametric, multivariate version of Analysis of Variance (ANOVA) that relies on a pseduo-F test to compare within-group and between group similarities based on a specified distance (dissimilarity) measure.
+Unlike ANOVA, PERMANOVA makes use of permutation tests to draw inferences without assuming distributions of test statistics.
+
+Second, distance matrices contain information on point-to-point distances (or dissimilarities) and, when vectorized, the correlation of the distance matrices measures similarity of two configurations (Mantel's test (Mantel, 1967) makes use of this approach to assess correlation between distance matrices).
+
+Third, correlations (or similarities) can be converted to distances; one such metric for doing this is $\sqrt{2(1-r_{ij})}$ (James et al, 2013).
+
+Finally, correlations can be computed using pairwise complete observations to leverage all available information to estimate correlations in the presence of some missing data values.
 
@@
 1.
-For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.
+For each observation in the dataset, remove it, and generate a map in 2 dimensions from the resulting (n-1) observations for a selected ordination method.
 (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA).
@@
 3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
-4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-|\sqrt{}\. 
+4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $$\sqrt{2(1-r_{ij})}$$ (James et al, 2013). 
 5.
@@
 
-We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Benjamini-Hochberg False Discovery Rate or Bonferroni corrections (or something similar).
+We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Benjamini-Hochberg False Discovery Rate or Bonferroni corrections (or something similar, assuming that the results are based on p-values as in the Bonferroni Outlier test (Cook & Weisberg, 1982)).
 We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth adonis test would be small.
@@
 
-```{r, eval = TRUE, message = FALSE}
+```{r, eval = TRUE, message = FALSE, warning = FALSE}
 #initializes some stuff 
+perplexity = 10
 data = yes_structure_final
@@
 
-3. Vectorize the distance matrix and generate a correlation matrix from the combination of these vectorized distances.
+3. Vectorize the distance matrix and generate a correlation matrix(Pearson correlation based on pairwise-complete observations) from the combination of these vectorized distances.
 
@@
 I'm ignoring the upper/lower triangle symmetry in this example.
-As an example, here's what this looks like in a very simplistic 3 observation case.
+As an example, here's what this looks like in a very simplistic 3 observation case, when the first observation is held out.
 
@@
 
-The vectorized form (shown as a row vector) would be: 
-$$\mathbf{c(Dist(X))^T} = \left[\begin{array}
+The vectorized form would be: 
+$$vec(\mathbf{c(Dist(X)))^T} = \left[\begin{array}
 {r}
@@
 $$
-Then, the correlation matrix as follows, where each column comes from the ith holdout map.
-Note that the below notation reads: $d_{23, 1}$ is the distance between observations 2 and 3 on the map that holds out the 1st observation.
+Then, the matrix used to compare distance matrices is as follows, where each column comes from the ith holdout map.
+Note that the below notation reads: $d_{23, -1}$ is the distance between observations 2 and 3 on the map that holds out the 1st observation.
 
@@
 {rrr}
-NA & d_{12, 2} & d_{13,3} \\
-NA & NA & d_{23,3} \\
-NA & d_{31,2} & NA \\
+NA & d_{12, -2} & d_{13,-3} \\
+NA & NA & d_{23,-3} \\
+NA & d_{31,-2} & NA \\
 NA & NA & ...\\
 0 & NA & .\\
-d_{32,1} & NA & .\\
-NA & d_{13,2} & .\\
-d_{23,1} & NA & .\\
+d_{32,-1} & NA & .\\
+NA & d_{13,-2} & .\\
+d_{23,-1} & NA & .\\
 0 & 0 & .\\
@@
 
-levelplot(as.matrix(distmat_list[[1]]), main = paste0("NA Pattern for ",1, "st Holdout"))
-levelplot(as.matrix(distmat_list[[3]]), main = paste0("NA Pattern for ",3, "rd Holdout"))
-levelplot(as.matrix(distmat_list[[5]]), main = paste0("NA Pattern for ",5, "th Holdout"))
+#levelplot(as.matrix(distmat_list[[1]]), main = paste0("NA Pattern for ",1, "st Holdout"))
+#levelplot(as.matrix(distmat_list[[3]]), main = paste0("NA Pattern for ",3, "rd Holdout"))
+#levelplot(as.matrix(distmat_list[[5]]), main = paste0("NA Pattern for ",5, "th Holdout"))
+
+image(as.matrix(distmat_list[[3]]), main = "NA Pattern for 3rd Holdout", col = hcl.colors(4))
 
@@
 The code chunk below shows both steps simultaneously.
-Note that generating the pearson correlation matrix from the vectorized distance matrices is done using pairwise complete observations - the NAs we include from the holdout set can cause potential problems when this is not specified.
 
@@
 
-  ## takes the vectorized distance matrices and smashes them into a single data frame - to build correlation matrix
+  ## takes the vectorized distance matrices and binds the columns into a single data frame - to build correlation matrix
 
@@
 4.
-Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using the result that $D_{ij} = \sqrt{2(1-cor)}$.
+Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using the result that $D_{-i,-j} = \sqrt{2(1-r_{-i,-j})}$.
 Note that here, we expect that all the correlation values are greater than 0 - and as shown, they are.
@@
 
-In the code below, the data does not change - the calls to 'adonis' change only in that the index of the "1" for the indicator variable chanegs at each iteration.
+In the code below, the data do not change - the calls to 'adonis' change only in that the index of the "1" for the indicator variable chanegs at each iteration.
 Thus, the indicator for the first holdout observation is [1,0,0,...] and so on and so forth.
 
-Note that for adonis, this version uses 1000 permutations rather than a permutation matrix, meaning that this version re-samples the $n+1$ different unique permutations of the indicator which has a single $1$ and $0$'s in all the other indices.
-*We are working on being a bit more efficient with this permutation strategy so that we do not sample more than necessary*.
+*Note that in some cases for Adonis, it may be possible to have more permutations than possible combinations of 0's and 1's - meaning that picking a number of permuations just re-samples the $n+1$ different unique permutations of the indicator which has a single $1$ and $0$'s in all the other indices (there are only* $n+1$ *pseudo-F statistics that are unique).
+To be more efficient with this permutation strategy so that we do not sample more than necessary, we specify a permutation matrix rather than passing a raw number of permuations*.
 
@@
 
-Note that in the Adonis function, we do obtain a few negative f-statistic values in certain cases.
+Note that in the Adonis function, we do obtain a few negative pseudo F-statistic values in certain cases.
 These are the null values based on the permutations used.
 This seems to occur only for the MDS data - in the cases where we do see negative values, they are quite close to 0.
+The sums of squares equivalent in PERMANOVA for the treatment is found by subtraction and the equivalent to the error sums of squares must be exceeding the total sums of squares (slightly, by a rounding error).
 
@@
 model_list[[1]]$f.perms %>% min()
-model_list[[1]]$f.perms %>% hist(main = "Example of Permutation Distribution: t-SNE")
+model_list[[1]]$f.perms %>% as.data.frame() %>% ggplot(aes(V1)) + geom_histogram(bins = 20) + ggtitle("Ex: Permutation Distribution: tSNE, bins = 20")
 
 #from cmd - an example with negatives included (note that they are all the same value )
+model_list_cmd[[1]]
 model_list_cmd[[1]]$f.perms %>% min()
 model_list_cmd[[1]]$f.perms[which(model_list_cmd[[1]]$f.perms<0)]
-model_list_cmd[[1]]$f.perms %>% hist(main = "Example of Permutation Distribution: MDS")
+model_list_cmd[[1]]$f.perms %>% as.data.frame() %>% ggplot(aes(V1)) + geom_histogram(bins = 20) + ggtitle("Ex: Permutation Distribution: MDS, bins = 20")
 

### 2021-09-21 PaperDraft.Rmd
adding updates to simulation docs

@@
 knitr::opts_chunk$set(echo = TRUE)
-source("OutlierCompFunctions.R")
+source("OutlierCompNew.R")
 library(ggplot2)
@@
 ## Sparse t-SNE
+
 ## Cross Validation and Mantel-esque Test
@@
 The idea is predicated on the concept that by leaving an out an influential observation, the results of the model should change.
-This process of jacknifing observations is done for all the observations in the data.
+This process of jackknifing observations is done for all the observations in the data.
 
@@
 
-```{r}
+```{r, eval = FALSE}
 #image of an influential point
@@
 
-The proposed approach is as follows:
+In general, we seek to identify either **multivariate outliers** (or groups of them) which, for some reason, cause large differences in the shape of the resulting ordination.
+The methods rely on a few different ideas.
+First, PERMANOVA (Anderson, 2001) is a non-parametric, multivariate version of Analysis of Variance (ANOVA) that relies on a pseduo-F test to compare within-group and between group similarities based on a specified distance (dissimilarity) measure.
+Unlike ANOVA, PERMANOVA makes use of permutation tests to draw inferences without assuming distributions of test statistics.
+
+Second, distance matrices contain information on point-to-point distances (or dissimilarities) and, when vectorized, the correlation of the distance matrices measures similarity of two configurations (Mantel's test (Mantel, 1967) makes use of this approach to assess correlation between distance matrices).
+
+Third, correlations (or similarities) can be converted to distances; one such metric for doing this is $\sqrt{2(1-r_{ij})}$ (James et al, 2013).
+
+Finally, correlations can be computed using pairwise complete observations to leverage all available information to estimate correlations in the presence of some missing data values.
+
+Our Permanova-based approach is as follows:
 
 1.
-For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.
+For each observation in the dataset, remove it, and generate a map in 2 dimensions from the resulting (n-1) observations for a selected ordination method.
 (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA).
@@
 3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
-4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-\|\sqrt{stuff}$. 
+4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $$\sqrt{2(1-r_{ij})}$$ (James et al, 2013). 
 5.
@@
 
-To test the method
+We conducted a series of simulations to assess the efficacy of the method at identifying influential points in various different situations.
+In all cases, we simulated data from multivariate normal distributions.
+Additionally, we did not use any of the data pre-processing available in any t-SNE maps; no PCA was used to pick an initial set of dimensions to reduce.
+
+In the simulation, each iteration simulates a new (contaminated) dataset, runs the t-SNE and MDS mapping procedure, and tests the consistency of the configurations for each of the *n* holdout points.
+Several versions of the simulations were assessed, including 1) 1 group with 3 "influental" points, 2) simulation with 3 groups with 3 "influential" points with different variances, and 3) simulation with 3 groups with 3 influential points with same variances.
+
+## Perplexity
+
+We also assessed the sensitivity of t-SNE's ability to identify influential points at various different levels of perplexity.
+
+## Sample Size and Dimensionality
+
+Two factors have the potential to greatly impact the computational complexity of generating maps: the dimensionality of the dataset and the sample size considered.
+For each of the mapping techniques examined, we assessed the
+
+The PERMANOVA step in the method can use either a pre-specified number of permutations, or it can be based on a permutation matrix that includes each of the permuted indices as rows.
+In small sample size situations, we utilize an exact p-value that is calculated using all of the possible permutations of the indicator variable as the explanatory variable consists only of a single 1 and the rest 0's.
+Thus, there are only $n$ distinct permutations to explore.
+In larger sample size scenarios, it may not be computationally feasible to consider every possible permutation of the indicator; instead, a raw number of permutations can be calculated and an estimated p-value can be calculated.
 

### 2021-09-21 Simulation Study Document V4.Rmd
adding updates to simulation docs

--- 2021-08-21 Simulation Study Document V3.Rmd
@@
 
-source("OutlierCompFunctions.R")
+source("OutlierCompNew.R")
 library(ggplot2)
@@
 
-```{r}
+```{r, echo = FALSE, eval = TRUE}
+
 n_sims = 100
@@
 results_list <- list()
+results_list_cmd <- list()
 df3list = df2list = list()
+
 ```
@@
 
-```{r original_sims, eval = FALSE, cache = TRUE, warning = FALSE, message = FALSE}
+Several versions of the simulations will be run:
+
++ 1 group with 3 "influental" points
++ 3 groups with 3 "influential" points with different variances
++ 3 groups with 3 influential points with same variances
+
++ Sample size increase/decrease (number of influential points relative to sample size too)
+
+## Simulation Chunks
+
+This chunk runs the simulation with slighlty increased variability in the three groups, and a restructuring of the influential points.
+Rather than simulating as outliers all on the positive side, I simulate one that has a large positive set of outlying points, one with a large negative set of outlying points, and one that simply 'bounces around' the multivariate space and breaks the grouping structure.
+
+```{r sims_v2, eval = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
+perplexity = 10
 
@@
 
-  #### Simulate the Data
+  #### Simulate the Data with outliers and a single group (simulates all from one group with mean 40)
 
-  g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(25, 10)) %>%
+  g1 <- rmnorm(n = 10, mean = rep(40, 10), diag(45, 10)) %>%
     t() %>%
     data.frame()
-  g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(25, 20)) %>%
+  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(45, 20)) %>%
     t() %>%
     data.frame()
-  g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(25, 30)) %>%
+  g3 <- rmnorm(n = 10, mean = rep(40, 30), diag(45, 30)) %>%
     t() %>%
@@
 
-  ## adds outliers in the last 3 indices (simulating from normal based on 5 sd's from mean)
-  add_outliers <- rbind(rnorm(10, 5, .5), rnorm(10, 5, .5), rnorm(10, 5, .5)) %>% as_tibble()
+  ### adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
+  add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
   names(add_outliers) <- names(yes_structure_scale)
@@
 
-  #Visualize the data and store in a list
-  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In")
+  ##### Visualize the data and store in a list
+  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
   plot_data_list[[j]] <- datplot
 
-  #### Run the Permanova mapping
+  #### Run the Permanova mapping (for t-SNE)
 
-  x3 <- MultiPermanova(yes_structure_final[,1:10])
+  xTSNE <- MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE")
+  results_list[[j]] <- xTSNE
 
-  results_list[[j]] <- x3
-
-  df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval), Fstat = sapply(x3$tsnelist, pullFstat))
+  df2 = tibble(Holdout = 1:length(xTSNE$modellist), PValues = sapply(xTSNE$modellist, pullPval), Fstat = sapply(xTSNE$modellist, pullFstat))
   df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
@@
 
-  ### CMD Plots #########################################################
-  df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval), Fstat = sapply(x3$cmdlist, pullFstat))
-    df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)
-  df3list[[j]] = df3
+   #### Run the Permanova mapping (for MDS)
 
-  #Visualize the results and store in a list
-  sigplot2 <- ggplot(df3, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: P-Values From Adonis")
-  plot_mds_results_list[[j]] <- sigplot2
-
-  #Visualize the f stats and store in a list
-  sigplot2f <- ggplot(df3, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("CMD: F Stats From Adonis")
-  plot_mds_fstat_results_list[[j]] <- sigplot2f
-
-}
-```
-
-This chunk runs the simulation with slighlty increased variability in the three groups, and a restructuring of the influential points.
-Rather than simulating as outliers all on the positive side, I simulate one that has a large positive set of outlying points, one with a large negative set of outlying points, and one that simply 'bounces around' the multivariate space and breaks the grouping structure.
-
-```{r sims_v2, eval = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
-
-for(j in 1:n_sims){
-
-  #### Simulate the Data with outliers adn a single group (simulates all from one group)
-
-  g1 <- rmnorm(n = 10, mean = rep(20, 10), diag(45, 10)) %>%
-    t() %>%
-    data.frame()
-  g2 <- rmnorm(n = 10, mean = rep(50, 20), diag(45, 20)) %>%
-    t() %>%
-    data.frame()
-  g3 <- rmnorm(n = 10, mean = rep(80, 30), diag(45, 30)) %>%
-    t() %>%
-    data.frame()
-
-  yes_structure <- rbind(g1, g2, g3) %>%
-    data.frame() %>%
-    mutate(Group = c(rep("1", 10), rep("2", 20), rep("3", 30)))
-
-  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
-
-  ## adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
-  add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
-  names(add_outliers) <- names(yes_structure_scale)
-
-  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
-  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))
-
-  #Visualize the data and store in a list
-  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
-  plot_data_list[[j]] <- datplot
-
-  #### Run the Permanova mapping
-
-  x3 <- MultiPermanova(yes_structure_final[,1:10])
-
-  results_list[[j]] <- x3
-
-  df2 = tibble(Holdout = 1:length(x3$tsnelist), PValues = sapply(x3$tsnelist, pullPval), Fstat = sapply(x3$tsnelist, pullFstat))
-  df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
-  df2list[[j]] = df2
-
-  #### TSNE Plots########################################################
-  #Visualize the results and store in a list
-  sigplot <- ggplot(df2, aes(Holdout, PValues)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: P-Values From Adonis")
-  plot_tsne_results_list[[j]] <- sigplot
-
-  #Visualize the f stats and store in a list
-  sigplotf <- ggplot(df2, aes(Holdout, Fstat)) + geom_point(aes(color = Colors), size = 2) + geom_line(alpha = 0.5, color = "grey") + theme_bw() + ggtitle("t-SNE: F Stats From Adonis")
-  plot_tsne_fstat_results_list[[j]] <- sigplotf
+   xCMD <- MultiPermanova(yes_structure_final[,1:10], maptype = "MDS")
+  results_list_cmd[[j]] <- xCMD
 
   ### CMD Plots #########################################################
-  df3 = tibble(Holdout = 1:length(x3$cmdlist), PValues = sapply(x3$cmdlist, pullPval), Fstat = sapply(x3$cmdlist, pullFstat))
+
+  df3 = tibble(Holdout = 1:length(xCMD$modellist), PValues = sapply(xCMD$modellist, pullPval), Fstat = sapply(xCMD$modellist, pullFstat))
     df3$Colors <- ifelse(df3$PValues < 0.05, TRUE, FALSE)
@@
 
-## PlotList
+## Plot List
 ```{r}

### 2021-09-21 Simulation-Study-Document-V3.Rmd
adding updates to simulation docs

--- 2021-09-21 Simulation Study Document V4.Rmd
@@
 
-source("OutlierCompNew.R")
+source("OutlierCompFunctions.R")
 library(ggplot2)

### 2021-09-21 Test.Rmd
adding updates to simulation docs

---
title: "Python Methods for Skillz App"
author: "Paul Harmon"
date: "8/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reads in the python files that we'll use.

```{python}
import numpy as np
import pandas as pd

peopleSkillz = pd.read_csv("https://raw.githubusercontent.com/AtriumAI/InternalSkillzApp/Keely26/issue138/temp_data%20/peopleSkillz.csv?token=AFIT5TOPNWUQ7MJN2ZYIGO3BGKWAC")
roleSkillz = pd.read_csv("https://raw.githubusercontent.com/AtriumAI/InternalSkillzApp/Keely26/issue138/temp_data%20/roleSkillz.csv?token=AFIT5TJ4TKCRILP5FNP7LX3BGKWEC")

peopleSkillz.head()
roleSkillz.head()
```

```{python}
prankings = pd.read_csv("C:/Users/paulh/Downloads/people_rankings.csv")
roles_temp = pd.read_csv("C:/Users/paulh/Downloads/people_rankings.csv")

roles_temp.head()
```

```{python}
#skill: array of skills for a given person
#role: array of roles for a given person
def match_skillz(persondf, roledf): 
  temp = persondf >= role 
  pct = sum(temp)/temp.size
  return pct
```

```{python}
#Method to generate a comparison between two vectors

#temp = np.array([3,3,2])  np.array([3,4,3])
#sum(temp)/temp.size

#peopleSkillz.columns.values.tolist()[1:]
#roleSkillz.columns.values.tolist()[3:]

#peopleSkillz[peopleSkillz.columns.values.tolist()[:]]
peopleSkillz[peopleSkillz['full_name'] == "Weisbeck, Keely"]
```

```{python}
##need a function to take role skillz and people skillz and just return a single array

def return_consultant(full_name): 
   df = peopleSkillz[peopleSkillz['full_name'] == full_name].iloc[0,1:]
   return df

```

```{python}
return_consultant("Weisbeck, Keely")
return_consultant("Harmon, Paul")
```

```{python}

def return_role(role_name): 
   df = roleSkillz[roleSkillz['name'] == role_name].iloc[0,3:]
   return df

```

```{python}
return_role('Data Scientist')
```

```{python}

#skill: array of skills for a given person
#role: array of roles for a given person
def match_skillz(skill, role): 
  temp = skill >= role 
  pct = sum(temp)/temp.size
  return pct

```

```{python}
t1 = np.array([3,4,5])
t2 = np.array([4,2,2])
match_skillz(t1, t2)
```
```{python}
roleSkillz.iloc[0:3, 0:1]
```

```{python}

## Script
# pick a role
# pick a consultant

# run the comparison
personskill = return_consultant("Weisbeck, Keely") 
personskill
role = return_role('Data Scientist') 
role
temp = match_skillz(personskill, role)

print("Match Score is:\n", temp)

```

### 2021-09-21 OutlierCompNew.R

@@ -1,6 +1,27 @@
 ## Draft version of outlier comparison function agnostic to mapping technique
 
 
+# preliminary functions: 
+
+## Vectorize Dist: Function that takes a dataframe and returns a vectorized version of a distance matrix applied to that data matrix. 
+## This function is used for generating vectorized data when we step through the Mantel test. 
+
+vectorize_dist <- function(df) {
+  tmp <- dist(df)
+  vec <- c(tmp)
+  return(vec)}
+
+## Pull P-Val and Pull F stat return the p-values and f-statstics from the list of output MultiPermanova puts out. 
+
+pullPval <- function(x){temp <- x$aov.tab$`Pr(>F`[1]; return(temp)}
+
+pullFstat <- function(x){temp <- x$aov.tab$`F.Model`[1]; return(temp)}
+
+
+
+
+
+
 #generate maps on holdout data (jth iteration is held out)
 # this function should be able to operate on multiple mapping techniques or a single one 
 
@@ -119,15 +140,15 @@ MultiPermanova <- function(data, nperms = 5000, perplexity = 10, pca_options = F
 
 
 #quick testing
-tic = Sys.time()
-x3 = MultiPermanova(yes_structure_final[,1:10], maptype = "MDS")
-toc = Sys.time()
-
-tic - toc
-
-tic = Sys.time()
-x4 = MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE", perplexity = 20)
-toc = Sys.time()
+# tic = Sys.time()
+# x3 = MultiPermanova(yes_structure_final[,1:10], maptype = "MDS")
+# toc = Sys.time()
+# 
+# tic - toc
+# 
+# tic = Sys.time()
+# x4 = MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE", perplexity = 20)
+# toc = Sys.time()
 
 #both functions take about 2.5 minutes
 

### 2021-09-21 PLOS paper files/plos_paper.tex
adding Pdf for PLOS one with additional updates to sims

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Influence Diagnostics for High-Dimensional Ordination Techniques} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Paul Harmon\textsuperscript{1,2\Yinyang},
Mark Greenwood\textsuperscript{2\Yinyang},
Laura Hildreth \textsuperscript{2,3\textcurrency},
Name4 Surname\textsuperscript{2},
Name5 Surname\textsuperscript{2\ddag},
Name6 Surname\textsuperscript{2\ddag},
Name7 Surname\textsuperscript{1,2,3*},
with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Montana State University  Deparment of Mathematical Sciences, Montana State University, Bozeman, MT, USA
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* correspondingauthor@institute.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam Eq~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id.~\cite{bib2} Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\begin{eqnarray}
\label{eq:schemeP}
	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
\end{eqnarray}

\section*{Materials and methods}
\subsection*{Etiam eget sapien nibh}

% For figure citations, please use "Fig" instead of "Figure".
Nulla mi mi, Fig~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

% Place figure captions after the first paragraph in which they are cited.
\begin{figure}[!h]
\caption{{\bf Bold the figure title.}
Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
\label{fig1}
\end{figure}

% Results and Discussion can be combined.
\section*{Results}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

% Place tables after the first paragraph in which they are cited.
\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{table}

%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
\subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
\subsubsection*{3rd level heading} 
Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit. Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat.

\begin{enumerate}
	\item{react}
	\item{diffuse free particles}
	\item{increment time by dt and go to 1}
\end{enumerate}

\subsection*{Sed ac quam id nisi malesuada congue}

Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\begin{itemize}
	\item First bulleted item.
	\item Second bulleted item.
	\item Third bulleted item.
\end{itemize}

\section*{Discussion}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero~\cite{bib3}.

\section*{Conclusion}

CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.

Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat. Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna. Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed. Nam condimentum sem eget mollis euismod. Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex. Nunc est dui, sodales sed mauris nec, auctor sagittis leo. Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor. For more information, see \nameref{S1_Appendix}.

\section*{Supporting information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 File.}
\label{S1_File}
{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Video.}
\label{S1_Video}
{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Appendix.}
\label{S1_Appendix}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\section*{Acknowledgments}
Cras egestas velit mauris, eu mollis turpis pellentesque sit amet. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
% 
\begin{thebibliography}{10}

\bibitem{bib1}
Conant GC, Wolfe KH.
\newblock {{T}urning a hobby into a job: how duplicated genes find new
  functions}.
\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.

\bibitem{bib2}
Ohno S.
\newblock Evolution by gene duplication.
\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
  Springer-Verlag.; 1970.

\bibitem{bib3}
Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
  infection through a transposon insertion followed by a {D}uplication}.
\newblock PLoS Genet. 2011 Oct;7(10):e1002337.

\end{thebibliography}

\end{document}

### 2021-09-21 PaperDraft.Rmd
adding Pdf for PLOS one with additional updates to sims

@@
 Consider the problem of visualizing multivariate (or potentially high-dimensional) data.
-In many cases, it is necessary to generate a 2 or 3-D mapping of some higher-dimensional space, projecting the high-dimensional data into an ordination that can be printed on paper.
+In many cases, it is useful to generate a 2 or 3-D mapping of some higher-dimensional space, projecting the high-dimensional data into an ordination that can be directly visualized.
 There are many methods that exist to accomplish this task, ranging from tools such as classicial multidimensional scaling (MDS) to more modern tools like t-distributed Stochastic Neighbor Embedding.
@@
 
-# High-dimensional Data Visualization Tools
+# Tools for High-Dimensional Data Visualization
 
@@
 Classical multidimensional scaling (MDS) is closely related to PCA (an in some cases, equivalent) (Gower, 1966).
-MDS takes as input a distance or dissimilarity matrix representing the distance between pairs of observations - this distance need not simply be Euclidean.
-Classical multidimensional scaling returns maps in teh same metric as the original distance or dissimilarity matrix.
+MDS takes as input a distance or dissimilarity matrix representing the distance between pairs of observations.
+This distance does not have to simply be Euclidean (although it often is).
+Classical multidimensional scaling returns maps in the same metric as the original distance or dissimilarity matrix; in general, the more dissimilar two observations are in the high-dimensional space, the farther apart they should be in the resulting MDS ordination.
 
-## Sparse t-SNE
+Multidimensional scaling attempts to minimize a metric called "stress" which can be defined generally as:
+$$ S_M(z_1, z_2, ..., z_N) = \sum_{i \neq i'}(d_{i,i'} - \|z_i - z_{i'}\|)^2 $$ 
+Different versions the stress function exist for different variations of MDS, including methods like Sammon mapping and Classical MDS, which frames the problem from a similarity perspective:
 
-## Cross Validation and Mantel-esque Test
+$$ S_c(z_1, z_2, ..., z_N) = \sum_{i, i'} (s_{i,i'} - (z_i - \bar{z}_,z_{i'} - \bar{z}))^2$$ 
+Classical MDS can be expressed equivalently as Principal Component Analysis when distances are Euclidean (Elements, XXXX).
+
+## Additional Mapping Techniques
+
+Placeholder section in case we want to add another mapping technique in.
+
++ SNE
++ Sammon Mapping
++ Isomap
 
@@
 
-# Simulation
+# Simulation Studies
 
@@
 
-```{r}
+```{r, eval = FALSE}
 CC2015 <- read.csv("https://raw.githubusercontent.com/paulharmongj/Carnegie_SEM/master/data/CC2015data.csv")
@@
 
-t-SNE is not particularly sensitive to single points that are highly dissimilar to others in the dataset.
+The proposed method uses a similar framework as to the tools available in regression in order to identify potentially abberant data points that can impact the shape of maps.
+Critically, the choice of ordination technique can impact the efficacy and sensitivity of maps to outliers/influential points.
+Some methods, like MDS, are highly responsive to influential points that do not look like the rest of the data.
+Maps resulting from these methods will typically make such points obvious.
+
+Alternatively, t-SNE is not particularly sensitive to single points that are highly dissimilar to others in the dataset.
 It is often not obvious that such observations are present when viewing a t-SNE map.
-In the case of other methods, like MDS, influential points change the configuration of the ordination significantly.
+In many cases, the influential data points are pushed towards the center of the resulting ordination, making them difficult to identify.
+
+Depending on the use case and problem to be solved, this sensitivity (or lack therof) may be benificial or potentially problematic.
+In some instances, it may be necessary to create maps that are relatively stable even in the presence of influential points; the results in this paper suggest that t-SNE might be more effective for creating maps in those situations.
+However, care should be taken when interpreting results from such maps as there may be substantively different observations that get grouped together in the resulting ordination.
+
+Much like in regression, we recommend taking an iterative approach to dealing with influential points.
+It is common practice in regression diagnostics to identify the point with the highest Cook's distance, remove it, and re-fit the model to see the new dispersion of observations' impacts on the model.
+A similar approach might be useful when creating high-dimensional data maps using these techniques.
+
+We recommend identifying the observation with the largest pseudo-F statistic, removing it, and re-assessing the PERMANOVA results with the reduced set.
+By cleaning out some of the most influential points, the resulting maps should be more reproducible, of higher fidelity, and provide potentially more meaningful results.

### 2021-09-21 Simulation with Single Profile.Rmd
adding Pdf for PLOS one with additional updates to sims

--- 2021-09-21 Simulation Study Document V4.Rmd
@@
 ---
-title: "Simulation Study Document V4"
+title: "Simulation Study with Feature Contribution"
 author: "Paul Harmon"
-date: "`r format(Sys.time(), '%B %d, %Y')`"
+date: "9/20/2021"
 output: html_document
@@
 
-This doc contains simulation studies around the efficacy of a Permanova-based method for identifying influential points in a t-SNE or MDS map.
-The document steps through a process of simulating data from 3 different groups (currently, these groups are clearly separated) with *3* influential points simulated on a standardized scale 5-6 standard deviations above the observed data.
-
-The code below outlines the number of simulations run - currently this number is small for computational ease but will be increased (with the use of adaptive Permanova methods).
+This doc looks at a profile where we have aberrant values only on a few of the features. Can we detect?
 
@@
 
-The below chunk runs the simulations.
-Each iteration simulates a new (contaminated) dataset, runs the t-SNE and MDS mapping procedure, and tests the consistency of the configurations for each of the *n* holdout points.
-**This code is not run in this version.**
+## Profile
 
-Several versions of the simulations will be run:
-
-+ 1 group with 3 "influental" points
-+ 3 groups with 3 "influential" points with different variances
-+ 3 groups with 3 influential points with same variances
-
-+ Sample size increase/decrease (number of influential points relative to sample size too)
-
-## Simulation Chunks
-
-This chunk runs the simulation with slighlty increased variability in the three groups, and a restructuring of the influential points.
-Rather than simulating as outliers all on the positive side, I simulate one that has a large positive set of outlying points, one with a large negative set of outlying points, and one that simply 'bounces around' the multivariate space and breaks the grouping structure.
+Let's look at a couple example profiles - here we might have a feature with only a few weird values.
 
 ```{r sims_v2, eval = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
-perplexity = 10
-
-for(j in 1:n_sims){
 
@@
 
-  g1 <- rmnorm(n = 10, mean = rep(40, 10), diag(45, 10)) %>%
+  g1 <- rmnorm(n = 20, mean = rep(40, 10), diag(45, 10)) %>%
     t() %>%
     data.frame()
-  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(45, 20)) %>%
+  g2 <- rmnorm(n = 20, mean = rep(40, 20), diag(45, 20)) %>%
     t() %>%
     data.frame()
-  g3 <- rmnorm(n = 10, mean = rep(40, 30), diag(45, 30)) %>%
+  g3 <- rmnorm(n = 20, mean = rep(40, 30), diag(45, 30)) %>%
     t() %>%
@@
 
-  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
+  yes_structure_scale <- lapply(yes_structure[,1:20], scale) %>% as_tibble()
 
-  ### adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
-  add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
+  ### adds outlier - where it's roughly the same for all but 2 of the features
+  add_outliers <- c(rnorm(18, 0, 1), c(-5,5))
   names(add_outliers) <- names(yes_structure_scale)
@@
   yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
-  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))
+  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
 
   ##### Visualize the data and store in a list
-  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
-  plot_data_list[[j]] <- datplot
+  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:20) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
+  datplot
 
+```
+
+```{r}
+perplexity = 10
   #### Run the Permanova mapping (for t-SNE)
 
-  xTSNE <- MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE")
+  xTSNE <- MultiPermanova(yes_structure_final[,1:20], maptype = "tSNE")
   results_list[[j]] <- xTSNE
@@
 }
-```
-
-# Results {.tabset}
-
-The simulated datasets can be seen in the tabset below.
-
-## Plot List
-```{r}
-library(ggpubr)
-#ggarrange(plotlist = plot_data_list)
-print(plot_data_list)
-```
-
-## Summary Metrics
-
-We can start to look at some results below.
-The code generates the rate of correctly classified influential points for both the MDS (shown first) and t-SNE maps.
-The MDS correctly classifies the influential point *some* of the time, but the t-SNE method very rarely classifies the influential points as having an impact on the configuration.
-
-```{r}
-
-#print the head of the df lists
-df3list %>% head(3)
-### Identify the number of CORRECTLY Classified influential points
-#extract the p-values from the last 3 from MDS
-influential_correct <- function(temp){length(which(tail(temp, 3)$PValues < 0.05))/3}
-mds_correct <- lapply(df3list, influential_correct)
-
-# extract the pvalues from the last 3 of the t-SNE
-tsne_correct <- lapply(df2list, influential_correct)
-
-dat = tibble(MDS = unlist(mds_correct), TSNE = unlist(tsne_correct))
-#plot(dat$MDS, dat$TSNE)
-
-mds_correct %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified")
-tsne_correct %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified")
-
-```
-
-Now we can look at the rate of misclassified non-influential points.
-These are the 60 'regular' points that are spuriously detected as outliers.
-Each percentage is calculated as a percentage of the 60 that are NOT small p-values (currenlty < 0.05 without a correction).
-We want a high percentage of correctly-identified non-influential points (although given the machinery of the hypothesis tests used, we would expect at least some spurious detections).
-
-Both are relatively high percentages (as we would expect) - but the t-SNE hovers right around the expected 0.95 and MDS is higher.
-
-```{r}
-
-### Identify rate of CORRECTLY Classified non-influential points
-#extract the p-values from the last 3 from MDS
-ni_correct <- function(temp){length(which(head(temp, 60)$PValues >= 0.05))/60}
-
-correct_mds <- lapply(df3list, ni_correct)
-#kable(correct_mds, col.names = "Correctly Identified Non-Influential: MDS")
-
-# extract the pvalues from the last 3 of the t-SNE
-correct_tsne <- lapply(df2list, ni_correct)
-#kable(correct_tsne, col.names = "Correctly Identified Non-Influential: tSNE")
-
-correct_mds %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified Non-Influential")
-correct_tsne %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified Non-Influential")
-
-```
-
-# Plots {.tabset}
-
-## MDS P Values
-
-Plots are shown below for the MDS .
-
-```{r}
-plot_mds_results_list
-```
-
-## MDS F Stats
-
-```{r}
-plot_mds_fstat_results_list
-```
-
-## tSNE
-
-```{r}
-plot_tsne_results_list
-```
-
-## t-SNE F Stats
-
-```{r}
-plot_tsne_fstat_results_list
-```

### 2021-09-30 Simulation with Single Profile.Rmd
Adding updated paper drafts

@@
 perplexity = 10
+
+ xTSNE <- MultiPermanova(yes_structure_final[,1:20], maptype = "tSNE")
+#  results_list[[j]] <- xTSNE
+
+  df2 = tibble(Holdout = 1:length(xTSNE$modellist), PValues = sapply(xTSNE$modellist, pullPval), Fstat = sapply(xTSNE$modellist, pullFstat))
+  df2$Colors <- ifelse(df2$PValues < 0.05, TRUE, FALSE)
+  df2list[[j]] = df2
+```
+
+```{r}
+perplexity = 10
   #### Run the Permanova mapping (for t-SNE)
@@
   plot_mds_fstat_results_list[[j]] <- sigplot2f
-
-}

### 2021-09-30 OutlierCompNew.R
Made improvements to the Outlier Comp New functions to better track perplexity

@@ -25,7 +25,7 @@ pullFstat <- function(x){temp <- x$aov.tab$`F.Model`[1]; return(temp)}
 #generate maps on holdout data (jth iteration is held out)
 # this function should be able to operate on multiple mapping techniques or a single one 
 
-Generate_Holdout_Maps <- function(data, j, maptype = "MDS",...){  
+Generate_Holdout_Maps <- function(data, j, maptype = "MDS", perp_val = 10){  
   
   #hold out a single row of the data (in the jth row)
   #note - have to remove, as t-SNE doesn't like NAs
@@ -42,7 +42,7 @@ Generate_Holdout_Maps <- function(data, j, maptype = "MDS",...){
   }
   else if(maptype == "tSNE"){
     #generate the t-SNE map and store x,y in map list
-    map <- Rtsne(data_dist, perplexity = perplexity, pca = pca_options, is_distance = TRUE)$Y
+    map <- Rtsne(data_dist, perplexity = perp_val, pca = pca_options, is_distance = TRUE)$Y
   }
   else{print("No maptype selected.")}
   
@@ -55,7 +55,7 @@ Generate_Holdout_Maps <- function(data, j, maptype = "MDS",...){
 
 #mapping method supported
 
-MultiPermanova <- function(data, nperms = 5000, perplexity = 10, pca_options = FALSE, maptype = "MDS"){
+MultiPermanova <- function(data, nperms = 1000, perp_val= 10, pca_options = FALSE, maptype = "MDS"){
   
   # initialize the size of the data and output lists
   n <- nrow(data)
@@ -140,15 +140,17 @@ MultiPermanova <- function(data, nperms = 5000, perplexity = 10, pca_options = F
 
 
 #quick testing
-# tic = Sys.time()
-# x3 = MultiPermanova(yes_structure_final[,1:10], maptype = "MDS")
-# toc = Sys.time()
+ tic = Sys.time()
+ x3 = MultiPermanova(yes_structure_final[,1:10], maptype = "MDS")
+ toc = Sys.time()
 # 
-# tic - toc
+ toc - tic
 # 
-# tic = Sys.time()
-# x4 = MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE", perplexity = 20)
-# toc = Sys.time()
+ tic = Sys.time()
+ x4 = MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE", perp_val = 20)
+ toc = Sys.time()
+ 
+ toc - tic
 
 #both functions take about 2.5 minutes
 

### 2021-09-30 Simulation Study Document with same mean same var.Rmd
changes to the doc

--- 2021-09-21 Simulation Study Document V4.Rmd
@@
 ---
-title: "Simulation Study Document V4"
+title: "Simulation Study Document: Same Mean Same Variance"
 author: "Paul Harmon"
@@
 
-  xTSNE <- MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE")
+  xTSNE <- MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE", perp_val = 20)
   results_list[[j]] <- xTSNE

### 2021-09-30 OutlierCompNew.R

@@ -140,17 +140,17 @@ MultiPermanova <- function(data, nperms = 1000, perp_val= 10, pca_options = FALS
 
 
 #quick testing
- tic = Sys.time()
- x3 = MultiPermanova(yes_structure_final[,1:10], maptype = "MDS")
- toc = Sys.time()
-# 
- toc - tic
-# 
- tic = Sys.time()
- x4 = MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE", perp_val = 20)
- toc = Sys.time()
- 
- toc - tic
+#  tic = Sys.time()
+#  x3 = MultiPermanova(yes_structure_final[,1:10], maptype = "MDS")
+#  toc = Sys.time()
+# # 
+#  toc - tic
+# # 
+#  tic = Sys.time()
+#  x4 = MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE", perp_val = 20)
+#  toc = Sys.time()
+#  
+#  toc - tic
 
 #both functions take about 2.5 minutes
 
